---
title: "Case study on B memory cell for improvement by consensus labels"
date: 15 Apr 2026
title-block-banner: true
author: 
  - name: Yingnan Gao
    orcid: 0000-0002-0960-4519
    email: yingnan.gao@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
        
  - name: Stefano Mangiola
    orcid: 0000-0001-7474-836X
    email: stefano.mangiola@adelaide.edu.au
    affiliation:
      - ref: sgc
      - id: weh
        name: Walter and Eliza Hall Institute
format: 
  html:
    theme: minty
    df-print: paged
    code-fold: true
    code-summary: "Show the code"
    code-line-numbers: true
    embed-resources: true
knitr:
  opts_chunk: 
    message: false
    warning: false
    echo: false
comments:
  hypothesis: 
    theme: clean
editor: source
---

## Identifying candidate conditions for showcasing improvement

To showcase an example of improvement by having refined or corrected cell type labels via ensemble consensus, I downloaded the cellNexus metadata joined by the original CELLxGENE metadata, and read the CELLxGENE unified cell type label dictionary deposited on GitHub.

```{r}
#| echo: true
#| results: hide
#| warning: false
#| message: false
#| fig-show: hide
library(cellNexus)
library(tidyverse)

# load mappings between cellxgene annotations and our unified cell types
cellxgene_map = read.csv("/Users/a1239028/Documents/GitHub/HPCell/inst/extdata/immune_map_cellxgene.csv")

# harmonise to the CL nomenclature
cellxgene_map = cellxgene_map |>
  select(from, to) |>
  rename(
    cell_type = from,
    unified_celltype = to
  )

# load cellNexus metadata and only keep high quality cells
### This will download ~1.5 GB of data if not already cached
cellNexus_metadata = get_metadata() |> 
  join_census_table() |> 
  keep_quality_cells()

```

I tracked the annotation change for cells originally labeled as B cell or B cell subtypes
in the original CELLxGENE unified cell types through the river plot.

```{r}
#| echo: true
#| results: hide
#| warning: false
#| message: false

b_mem_alluvial_data = cellNexus_metadata |>
  left_join(cellxgene_map,copy = TRUE)|>
  filter(unified_celltype %in% c("b memory","b naive","plasma","b")) |>
  mutate(
    unified_celltype = case_when(
      unified_celltype %in% c("b memory","b naive","plasma","b") ~ unified_celltype,
      .default="Other"
    ),
    cell_type_unified_ensemble = case_when(
      cell_type_unified_ensemble %in% c("b memory","b naive","plasma","b") ~ cell_type_unified_ensemble,
      .default="Other"
    )
  ) |>
  count(unified_celltype,cell_type_unified_ensemble) |>
  mutate(
    original_level = case_when(
      unified_celltype %in% c("b memory","b naive","plasma") ~ "L2",
      .default= "L1"
    ),
    consensus_level = case_when(
      cell_type_unified_ensemble %in% c("b memory","b naive","plasma") ~ "L2",
      .default= "L1"
    )
  ) |>
  collect()

b_mem_alluvial_data$cell_type_unified_ensemble = 
  factor(b_mem_alluvial_data$cell_type_unified_ensemble,
         levels=c("Other","b","b memory","b naive","plasma"))

b_mem_alluvial_data$unified_celltype = 
  factor(b_mem_alluvial_data$unified_celltype,
         levels=c("Other","b","b memory","b naive","plasma"))

library(ggalluvial)

# Immune graph level color scheme
immune_level_color = c(
  L0 = "#001959",
  L1 = "#3B6C55",
  L2 = "#D29243",
  L3 = "#F9CCF9"
)

celltype_level = c("b memory"="L1","b naive"="L1","plasma"="L1","b"="L0","Other"="L0")
celltype_color = immune_level_color[celltype_level]
names(celltype_color) = names(celltype_level)

b_mem_alluvial = 
  ggplot(b_mem_alluvial_data,
         aes(y = n, axis1 = unified_celltype, axis2 = cell_type_unified_ensemble)) +
    geom_alluvium(aes(fill = original_level), width = 0.3,alpha=0.75) +
    geom_stratum(mapping = aes(fill=after_stat(stratum)),width = 0.3) +
    geom_text(stat = "stratum", mapping = aes(label = after_stat(stratum)),color="white") +
    scale_x_discrete(limits = c("unified_celltype", "cell_type_unified_ensemble"), expand = c(.05, .05),
                     labels=c("unified_celltype"="original CELLxGENE",
                              "cell_type_unified_ensemble" = "ensemble consensus")) +
    scale_fill_manual(values = c(immune_level_color,celltype_color),name="Level")+
    scale_y_continuous(breaks=c(0,5e5,1e6,1.5e6,2e6),
                       labels=sapply(
                         c("0","5%*%10^5","1%*%10^6","1.5%*%10^6","2%*%10^6"),
                         function(x){parse(text=x)}))+
    ylab("Cell count")+
    theme_minimal()+
    theme(legend.position = "none")

print(b_mem_alluvial)
```

It appears that B memory is a good candidate to as it has the highest degree of improvement.

I then counted the number of cells that are labelled as B memory cells within each tissue group and disease condition, and the number of samples that contains these cells within the same tissue group and disease condition, for both the original CELLxGENE annotation and the ensemble consensus annotation.

```{r}
#| echo: true
#| results: hide
#| warning: false
#| message: false
#| fig-show: hide
# a count on samples and cells by tissue and disease condition
original_disease_tissue_sample_count = cellNexus_metadata|>
  left_join(cellxgene_map,copy = TRUE)|>
  filter(unified_celltype %in% c("b memory","b"),
         grepl("memory",x = cell_type)) |>
  count(tissue_groups,sample_id,disease)|>
  collect()

original_disease_tissue_count = left_join(
  x = original_disease_tissue_sample_count|>
    count(tissue_groups,disease,wt=n)|>
    rename(n_cell=n),
  y = original_disease_tissue_sample_count|>
    count(tissue_groups,disease)|>
    rename(n_sample=n)
)

original_disease_tissue_count_by_disease = split(
  original_disease_tissue_count,original_disease_tissue_count$disease)

original_disease_tissue_count_by_tissue = split(
  original_disease_tissue_count,original_disease_tissue_count$tissue_groups)

consensus_disease_tissue_sample_count = cellNexus_metadata|>
  left_join(cellxgene_map,copy = TRUE)|>
  filter(cell_type_unified_ensemble %in% c("b memory")) |>
  count(tissue_groups,sample_id,disease)|>
  collect()

consensus_disease_tissue_count = left_join(
  x = consensus_disease_tissue_sample_count|>
    count(tissue_groups,disease,wt=n)|>
    rename(n_cell=n),
  y = consensus_disease_tissue_sample_count|>
    count(tissue_groups,disease)|>
    rename(n_sample=n)
)

consensus_disease_tissue_count_by_disease = split(
  consensus_disease_tissue_count,consensus_disease_tissue_count$disease)

consensus_disease_tissue_count_by_tissue = split(
  consensus_disease_tissue_count,consensus_disease_tissue_count$tissue_groups)
```

For healthy samples, I generated a table for tissues that has at least one sample containing B memory cells in both original and ensemble consensus annotation.

```{r}
#| echo: true
#| results: hide
#| warning: false
#| message: false
#| fig-show: hide
# find candidates for DE between tissue in healthy samples 
healthy_by_tissue_comparison = left_join(
  original_disease_tissue_count_by_disease$normal |>
    select(-disease)|> rename(n_cell_original=n_cell,n_sample_original = n_sample),
  consensus_disease_tissue_count_by_disease$normal|>
    select(-disease)|> rename(n_cell_consensus=n_cell,n_sample_consensus = n_sample),
)
```

```{r}
#| echo: false
healthy_by_tissue_comparison |> select(-n_cell_original,-n_cell_consensus) |> 
  print(n = Inf)
```

I also generated a table for disease conditions to find potential comparison between healthy samples and disease samples.

```{r}
#| echo: true
#| results: hide
#| warning: false
#| message: false
#| fig-show: hide
# find candidates for DE between healthy and disease in each tissue
healthy_and_disease_comparison = do.call(rbind,lapply(
  intersect(names(original_disease_tissue_count_by_tissue),
            names(consensus_disease_tissue_count_by_tissue)),
  function(nn){
    df = left_join(
      original_disease_tissue_count_by_tissue[[nn]] |>
        rename(n_cell_original=n_cell,n_sample_original = n_sample),
      consensus_disease_tissue_count_by_tissue[[nn]]|>
        rename(n_cell_consensus=n_cell,n_sample_consensus = n_sample),
    )
    if(dim(df)[1]<2) return(NULL)
    return(df)
  }))
```

```{r}
#| echo: false
healthy_and_disease_comparison|>select(-n_cell_original,-n_cell_consensus) |> 
  print(n = Inf)
```

It appears that comparison between blood and fat (adipose) tissues will benefit the most from having the ensemble consensus annotation.

## Pseudo-bulk analysis using original and ensemble consensus annotation

I then conducted pseudo-bulk DE analysis using the R package `tidybulk`. 

To factor out potential batch effects by covariates like datasets and sequencing
technology, I updated the cellNexus metadata based on a deposited function:
(https://gist.github.com/stemangiola/8fe6c45b79dd95a200c0fd2314ec57d0).

```{r}
#| echo: true
#| fig-show: hide

assay_data_grouped = tribble(
  ~assay, ~assay_groups,
  "10x 3' v2", "10x Genomics 3",
  "10x 3' v3", "10x Genomics 3",
  "10x 5' v2", "10x Genomics 5",
  "10x 5' v1", "10x Genomics 5",
  "MARS-seq", "Plate based Technologies",
  "10x 3' transcription profiling", "10x Genomics 3",
  "10x 5' transcription profiling", "10x Genomics 5",
  "Smart-seq2", "Smart seq",
  "microwell-seq", "Microwell Technologies",
  "TruDrop", "TruDrop",
  "Drop-seq", "Drop based Technologies",
  "Seq-Well S3", "Microwell Technologies",
  "GEXSCOPE technology", "Other Technologies",
  "Seq-Well", "Microwell Technologies",
  "sci-RNA-seq", "Other Technologies",
  "10x 3' v1", "10x Genomics 3",
  "BD Rhapsody Whole Transcriptome Analysis", "Other Technologies",
  "BD Rhapsody Targeted mRNA", "Other Technologies",
  "CEL-seq2", "Plate based Technologies",
  "SPLiT-seq", "Other Technologies",
  "STRT-seq", "Plate based Technologies",
  "inDrop", "Drop based Technologies",
  "Smart-seq v4", "Smart seq",
  "ScaleBio single cell RNA sequencing", "Other Technologies"
)

cellNexus_metadata = cellNexus_metadata |>
  left_join(assay_data_grouped, copy=TRUE)

```

### Case 1: healthy fat vs healthy blood

Because I need pseduo-bulk data aggregated under the original CELLxGENE annotation, I pulled the relevant single-cell data and aggregated them manually for both annotations.

```{r}
#| echo: true
#| fig-show: hide
library(tidybulk)
library(edgeR)
library(limma)
library(variancePartition)
library(patchwork)

# collect metadata
original_healthy_fat = cellNexus_metadata |>
  left_join(cellxgene_map,copy = TRUE)|>
  filter(unified_celltype %in% c("b memory","b"),
         grepl("memory",x = cell_type),
         disease=="normal",
         feature_count > 5000,
         tissue_groups%in%c("adipose tissue")) |>
  collect()

original_fat_datasets = unique(original_healthy_fat$dataset_id)

original_healthy_blood = cellNexus_metadata |>
  left_join(cellxgene_map,copy = TRUE)|>
  filter(unified_celltype %in% c("b memory","b"),
         grepl("memory",x = cell_type),
         disease=="normal",
         feature_count > 5000,
         dataset_id %in% !!original_fat_datasets,
         tissue_groups %in% c("blood")) |>
  collect()

consensus_healthy_fat = cellNexus_metadata |>
  left_join(cellxgene_map,copy = TRUE)|>
  filter(cell_type_unified_ensemble %in% c("b memory"),
         disease=="normal",
         feature_count > 5000,
         #dataset_id %in% !!original_fat_datasets,
         assay_groups %in% c("10x Genomics 5","10x Genomics 3"),
         tissue_groups%in%c("adipose tissue")) |>
  collect()

consensus_fat_datasets = unique(consensus_healthy_fat$dataset_id)

consensus_healthy_blood = cellNexus_metadata |>
  left_join(cellxgene_map,copy = TRUE)|>
  filter(cell_type_unified_ensemble %in% c("b memory"),
         disease=="normal",
         feature_count > 5000,
         dataset_id %in% !!consensus_fat_datasets,
         tissue_groups %in% c("blood")) |>
  collect()

original_healthy_fat_vs_blood_SCE = rbind(original_healthy_fat,original_healthy_blood)|>
  get_single_cell_experiment()

consensus_healthy_fat_vs_blood_SCE = rbind(consensus_healthy_fat,consensus_healthy_blood)|>
  get_single_cell_experiment()

```

I filtered the aggregated pseudo-bulk data to keep only the features that are abundant
across 70% of the sample in at least one tissue, and then normalized the gene expression
counts with method 'TMMwsp'.

```{r}
#| echo: true
#| fig-show: hide

# aggregate from single cell data
as_SE = function(x){
  rn = rownames(x)
  x = as(x,"RangedSummarizedExperiment")
  rownames(x) = rn
  #colnames(x) = paste0(x@colData$sample_id,"___",x@colData$tissue_groups)
  return(x)
}

original_healthy_fat_vs_blood_SE = original_healthy_fat_vs_blood_SCE |>
  scuttle::aggregateAcrossCells(
    ids = paste0(original_healthy_fat_vs_blood_SCE@colData$tissue_groups,
                 "_",original_healthy_fat_vs_blood_SCE@colData$donor_id),
    statistics = "sum"
  ) |> as_SE() |>
  keep_abundant_per_category(formula_design = ~ tissue_groups,
                minimum_count_per_million = 10,
                minimum_proportion = 0.7) |>
  scale_abundance(method = "TMMwsp") #|>
  #resolve_complete_confounders_of_non_interest(sex,dataset_id,assay_groups)

print(dim(original_healthy_fat_vs_blood_SE))

consensus_healthy_fat_vs_blood_SE = consensus_healthy_fat_vs_blood_SCE |>
  scuttle::aggregateAcrossCells(
    ids = paste0(consensus_healthy_fat_vs_blood_SCE@colData$tissue_groups,
                 "_",consensus_healthy_fat_vs_blood_SCE@colData$donor_id),
    statistics = "sum"
  ) |> as_SE() |>
  keep_abundant_per_category(formula_design = ~ tissue_groups,
                minimum_count_per_million = 10,
                minimum_proportion = 0.7) |>
  scale_abundance(method = "TMMwsp") #|>
  #resolve_complete_confounders_of_non_interest(sex,dataset_id,assay_groups)

print(dim(consensus_healthy_fat_vs_blood_SE))

```

Then, I conducted differential expression analysis through 'dream()' with a focus on
coefficients of tissue groups in pseudobulk data aggregated based on original CELLxGENE
annotation:

```{r}
#| echo: true

# simplify dataset id and tissue group for contrast names
simplify_complex_entry_in_colData = function(se,levels){
  se@colData$tissue_groups_simplified = 
    paste0("tissue_",as.numeric(factor(se@colData$tissue_groups,levels = levels)))
  if(is.null(se@colData$dataset_id___altered)){
    se@colData$dataset_id_simplified = 
      paste0("dataset_",as.numeric(factor(se@colData$dataset_id)))
  }else{
    se@colData$dataset_id_simplified = 
      paste0("dataset_",as.numeric(factor(se@colData$dataset_id___altered)))
  }
  if(is.null(se@colData$assay_groups___altered)){
    se@colData$assay_groups_simplified = 
      paste0("assay_",as.numeric(factor(se@colData$assay_groups)))
  }else{
    se@colData$assay_groups_simplified = 
      paste0("assay_",as.numeric(factor(se@colData$assay_groups___altered)))
  }
  return(se)
}

# DE using dream()
original_healthy_fat_vs_blood_SE = 
  original_healthy_fat_vs_blood_SE |>
  simplify_complex_entry_in_colData(levels=c("blood","adipose tissue"))

original_healthy_fat_vs_blood_formula = ~ tissue_groups_simplified + sex + (1|donor_id)
original_healthy_fat_vs_blood_dge = 
  original_healthy_fat_vs_blood_SE@assays@data$counts |>
  DGEList()|>
  calcNormFactors(method="TMMwsp")

original_healthy_fat_vs_blood_voom = 
  original_healthy_fat_vs_blood_dge |>
  voomWithDreamWeights(original_healthy_fat_vs_blood_formula,
                       original_healthy_fat_vs_blood_SE@colData,
                       plot = TRUE)

original_healthy_fat_vs_blood_fit = 
  original_healthy_fat_vs_blood_voom |>
  dream(original_healthy_fat_vs_blood_formula,
        original_healthy_fat_vs_blood_SE@colData) |>
  eBayes()

head(original_healthy_fat_vs_blood_fit$design,3)
```

and the ensemble consensus annotation:

```{r}
#| echo: true
consensus_healthy_fat_vs_blood_SE = 
  consensus_healthy_fat_vs_blood_SE |>
  simplify_complex_entry_in_colData(levels=c("blood","adipose tissue"))

consensus_healthy_fat_vs_blood_formula =
  ~ tissue_groups_simplified + sex + (1|donor_id) + (1|dataset_id_simplified)

consensus_healthy_fat_vs_blood_dge = 
  consensus_healthy_fat_vs_blood_SE@assays@data$counts |>
  DGEList()|>
  calcNormFactors(method="TMMwsp")

consensus_healthy_fat_vs_blood_voom = 
  consensus_healthy_fat_vs_blood_dge |>
  voomWithDreamWeights(consensus_healthy_fat_vs_blood_formula,
                       consensus_healthy_fat_vs_blood_SE@colData,
                       plot = TRUE)

consensus_healthy_fat_vs_blood_fit = 
  consensus_healthy_fat_vs_blood_voom |>
  dream(consensus_healthy_fat_vs_blood_formula,
        consensus_healthy_fat_vs_blood_SE@colData) |>
  eBayes()

head(consensus_healthy_fat_vs_blood_fit$design,3)

```

I plotted the distribution of counts per sample and p-values,

```{r}
#| echo: true
# Histogram of p-values
limma::plotDensities(log1p(original_healthy_fat_vs_blood_SE@assays@data$counts_scaled),
                     legend=FALSE,main="original")

limma::plotDensities(log1p(consensus_healthy_fat_vs_blood_SE@assays@data$counts_scaled),
                     legend=FALSE,main="consensus")

ori_fat_blood_pval_hist = 
  original_healthy_fat_vs_blood_fit |>
  topTable(number = Inf,coef = "tissue_groups_simplifiedtissue_2") |>
  mutate(
    annotation_method = factor("original CELLxGENE",
                               levels=c("original CELLxGENE","ensemble consensus")))

con_fat_blood_pval_hist = 
  consensus_healthy_fat_vs_blood_fit |>
  topTable(number = Inf,coef = "tissue_groups_simplifiedtissue_2") |>
  mutate(
    annotation_method = factor("ensemble consensus",
                               levels=c("original CELLxGENE","ensemble consensus")))

b_mem_pval_hist = 
  rbind(ori_fat_blood_pval_hist,con_fat_blood_pval_hist)|>
  ggplot() +
  geom_histogram(mapping = aes(x = P.Value),binwidth = 0.01,fill="darkgrey") +
  facet_wrap(~annotation_method,ncol=1,nrow=2) +
  labs(x="P-Value",y="Count")+
  theme_minimal() +
  theme(legend.position = "none")

print(b_mem_pval_hist)
```

and the LFC-mean plot for QC.

```{r}
#| echo: true
# LFC-mean
original_healthy_fat_vs_blood_fit |>
  plotMA(main="original")

consensus_healthy_fat_vs_blood_fit |>
  plotMA(main="consensus")
```

Then I selected the top ten up- and down-regulated genes in adipose tissues using both the
original and the consensus label:

```{r}
#| echo: true
#| fig-show: hide
library(org.Hs.eg.db)
ori_fat_blood_outcome = ori_fat_blood_pval_hist |>
  rownames_to_column(var=".feature") |>
  arrange(adj.P.Val) |> mutate(
  genename = mapIds(
    org.Hs.eg.db,
    keys = .feature,
    column = "GENENAME",
    keytype = "ENSEMBL",
    multiVals = "first"),
  symbol = mapIds(
    org.Hs.eg.db,
    keys = .feature,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first")
)

print("original label:")

ori_fat_blood_outcome |>   filter(logFC>0) |>
  head(10)  |> dplyr::select(genename,symbol,logFC,P.Value,adj.P.Val) |> as.tibble() |> print()

ori_fat_blood_outcome |>   filter(logFC<0) |>
  head(10)  |> dplyr::select(genename,symbol,logFC,P.Value,adj.P.Val) |> as.tibble() |> print()

con_fat_blood_outcome = con_fat_blood_pval_hist |>
  rownames_to_column(var=".feature") |>
  arrange(adj.P.Val) |> mutate(
  genename = mapIds(
    org.Hs.eg.db,
    keys = .feature,
    column = "GENENAME",
    keytype = "ENSEMBL",
    multiVals = "first"),
  symbol = mapIds(
    org.Hs.eg.db,
    keys = .feature,
    column = "SYMBOL",
    keytype = "ENSEMBL",
    multiVals = "first")
)

print("consensus label:")

con_fat_blood_outcome |>  filter(logFC>0) |>
  head(10)  |> dplyr::select(genename,symbol,logFC,P.Value,adj.P.Val) |> as.tibble() |> print()

con_fat_blood_outcome |>  filter(logFC<0) |>
  head(10)  |> dplyr::select(genename,symbol,logFC,P.Value,adj.P.Val) |> as.tibble() |> print()
```

I checked the volcano plot:

```{r}
#| echo: true
# Top residency gene candidates

original_healthy_fat_vs_blood_SE@colData |> as_tibble() |>
  count(tissue_groups,tissue_groups_simplified)

consensus_healthy_fat_vs_blood_SE@colData |> as_tibble() |>
  count(tissue_groups,tissue_groups_simplified)

fat_enrich_blood_genes = c(
"CXCR4","ITGA4","ITGB1","CD44","BCL2",
"HLA-DRA","CD69","CD74","NFKBIA","TNFAIP3"
)

# Create volcano plots
b_mem_volcano_data = 
  rbind(
    ori_fat_blood_outcome |>
    # Add candidate annotation
    mutate(expectation  = case_when(
        symbol %in% fat_enrich_blood_genes ~ "Enriched in adipose tissue",
        .default = "Not candidate"
        ),
        label = case_when(
          symbol %in% fat_enrich_blood_genes ~ symbol,
          .default = NA
        ),
        annotation_method = factor("original CELLxGENE",
                                   levels=c("original CELLxGENE","ensemble consensus"))
    ) |> arrange(desc(expectation)),
    con_fat_blood_outcome |>
    # Add candidate annotation
    mutate(expectation  = case_when(
        symbol %in% fat_enrich_blood_genes ~ "Enriched in adipose tissue",
        .default = "Not candidate"
        ),
        label = case_when(
          symbol %in% fat_enrich_blood_genes ~ symbol,
          .default = NA
        ),
        annotation_method = factor("ensemble consensus",
                                   levels=c("original CELLxGENE","ensemble consensus"))
    ) |> arrange(desc(expectation))
  )

# Plot
b_mem_volcano = b_mem_volcano_data |>
  ggplot(aes(x = logFC, y = P.Value)) +
  geom_point(aes(color = expectation, size = adj.P.Val < 0.05)) +
  geom_hline(yintercept = 0.05,color="black",linetype="dashed")+
  ggrepel::geom_text_repel(aes(label = label), size = 2, max.overlaps = 20) +
  scale_x_continuous(breaks=seq(-6,6,1))+
  scale_y_continuous(trans = "log10_reverse",
                     breaks = 10^seq(-9,0,1),
                     labels=sapply(seq(-9,0,1),function(x){
                       parse(text=sprintf("10^%d",x))
                       })) +
  scale_color_manual(values = c("Enriched in adipose tissue" = "red","Not candidate" ="grey")) +
  scale_size_manual(values = c("TRUE" = 0.75, "FALSE" = 0.1)) +
  facet_wrap(facets=vars(annotation_method),ncol=1,nrow=2,scales="free_y")+
  theme_minimal() +
  labs(x="log-FC",y="P-value")

print(b_mem_volcano)
```

The aggregated result figure for manuscript:

```{r}
#| echo: true
b_mem_alluvial + b_mem_pval_hist + b_mem_volcano + plot_layout(ncol = 3,nrow = 1)
```