---
title: "Build and annotate Human Cell Atlas Ontology immune graph"
date: 15 Apr 2026
title-block-banner: true
author: 
  - name: Yingnan Gao
    orcid: 0000-0002-0960-4519
    email: yingnan.gao@adelaide.edu.au
    affiliation:
      - id: sgc
        name: SAiGENCI, The University of Adelaide
        
  - name: Stefano Mangiola
    orcid: 0000-0001-7474-836X
    email: stefano.mangiola@adelaide.edu.au
    affiliation:
      - ref: sgc
      - id: weh
        name: Walter and Eliza Hall Institute
format: 
  html:
    theme: minty
    df-print: paged
    code-fold: true
    code-summary: "Show the code"
    code-line-numbers: true
    embed-resources: true
knitr:
  opts_chunk: 
    message: false
    warning: false
    echo: true
comments:
  hypothesis: 
    theme: clean
editor: source
---

# Programmatic building of immune graph from Human Cell Atlas Ontology (HCAO)

HCAO inherits Cell Ontology (CL) terms for cell types, and thus we use CL as the
standard cell type identifier across references.

## Downloading HCAO

The full HCAO can be downloaded as a .owl file and parsed by `rdflib` package in
R:

```{r}
library(rdflib)
library(igraph)
library(ggraph)

# read Human Cell Atlas Ontology, as of 20260330
if(!file.exists("hcao.owl")){
  download.file(
    url = "https://ontology.archive.data.humancellatlas.org/ontologies/hcao/download",
    destfile = "hcao.owl") 
  # md5sum: a1239028@LHDF2W0M6D0, as of 20260330
}
hcao = rdf_parse("hcao.owl") 
```

## Query HCAO for cell types

We query all term pairs with a `subClassOf` relationship, and filtered them to include
only those CL pairs:

```{r}
# set query
ontology.sparql = "
  PREFIX rdfs:<http://www.w3.org/2000/01/rdf-schema#>
  PREFIX oboInOwl:<http://www.geneontology.org/formats/oboInOwl#>
  PREFIX owl:  <http://www.w3.org/2002/07/owl#>
  PREFIX obo:  <http://purl.obolibrary.org/obo/>
  PREFIX skos:<http://www.w3.org/2004/02/skos/core#>
  SELECT ?child ?parent ?clab ?plab
  WHERE {
    ?child rdfs:subClassOf ?parent .
    ?child rdfs:label ?clab .
    ?parent rdfs:label ?plab}"

# Extract subclass relationships (edges) for only Cell Ontology (CL)
hcao.triples = rdf_query(hcao, ontology.sparql)
hcao.triples = hcao.triples[
  intersect(grep("http://purl.obolibrary.org/obo/CL",hcao.triples$parent),
            grep("http://purl.obolibrary.org/obo/CL",hcao.triples$child)),]

# simplify tag
hcao.df = hcao.triples
hcao.df$child = sapply(hcao.triples$child,function(x){rev(strsplit(x,"/")[[1]])[1]})
hcao.df$parent = sapply(hcao.triples$parent,function(x){rev(strsplit(x,"/")[[1]])[1]})
```

Some CL terms associate with multiple common name labels, and thus have multiple
entries in the table. We only include the shortest common name.

```{r}
# remove duplicate labels
cl_labels = split(c(hcao.df$clab,hcao.df$plab),c(hcao.df$child,hcao.df$parent))
unique_cl_labels = sapply(cl_labels,function(x){x[which.min(sapply(x,nchar))]})
hcao.unique.df = unique(hcao.df[,c("child","parent")])
```

## Building HCAO immune graph

We start from the canonical root of all immune cells in the ontology, CL_0000988
(hematopoietic cell), to recursively build the immune graph.

```{r, results='hide'}
# extract immune ontology tree
cl_root = "CL_0000988" # common root of immune cell (hematopoietic cell)
nodes_to_parse = cl_root
hcao_cl_rooted.df = hcao.unique.df[numeric(0),]
hcao_cl_rooted.df$depth = numeric(0)
hcao_depth = 1
while(length(nodes_to_parse)>0){
  idx = which((hcao.unique.df$parent%in%nodes_to_parse)&(!hcao.unique.df$parent%in%hcao_cl_rooted.df$parent))
  if(length(idx)<=0) break
  these_childs = unique(hcao.unique.df$child[idx])
  these_df = hcao.unique.df[idx,]
  these_df$depth = hcao_depth
  hcao_cl_rooted.df = rbind(hcao_cl_rooted.df,these_df)
  nodes_to_parse = these_childs
  hcao_depth = hcao_depth+1
}

# use the shortest label for each ontology term
hcao_cl_rooted.df$clab=unique_cl_labels[hcao_cl_rooted.df$child]
hcao_cl_rooted.df$plab=unique_cl_labels[hcao_cl_rooted.df$parent]

# build igraph object for ensemble functions
hcao.dag = graph_from_data_frame(d = hcao_cl_rooted.df[, c("child", "parent")],
                                 directed = TRUE)
V(hcao.dag)$common_label = unique_cl_labels[V(hcao.dag)$name]
V(hcao.dag)$common_label[is.na(V(hcao.dag)$common_label)] = "non_immune"

# clean up space
rm(hcao)
gc()
```

## Building reference mapping dictionary

In addition to the full HCAO immune graph, we build the dictionary between references
and HCAO. For BluePrint, we use `celldex` to associate reference annotation
labels and CL terms.

```{r, results='hide'}
### Making label to CL/HCAO tree dictionary

# Get blueprint metadata via celldex
blueprint.data = celldex::BlueprintEncodeData()
blueprint.labels = split(f = blueprint.data$label.fine, x = blueprint.data$label.ont)
# check if fine-level label to ontology is one to one
print(table(sapply(blueprint.labels,function(x){length(unique(x))})))
# make dictionary
blueprint_label_fine_to_ont = do.call(
  rbind,
  lapply(names(blueprint.labels),function(x){
    CL = gsub(pattern = ":",replacement = "_",x = blueprint.labels[[x]][1],fixed = TRUE)
    df = data.frame(from=x,to=CL,is_immune=TRUE)
    if(!df$to%in%hcao_cl_rooted.df$child) df$to = "not hematopoietic"
    return(df)
  })
)
# save to file
write.table(blueprint_label_fine_to_ont,file = "immune_map_blueprint_to_CL.csv",
            quote = FALSE,row.names = FALSE,col.names = TRUE,sep=",")
#
rm(blueprint.data)
gc()

```

For Monaco, we also use celldex but we manually curate one annotation of Th1/Th17
to Th17 as the celldex CL annotation is isolated from Th1 or Th17 cells in HCAO
immune graph.

```{r, results='hide'}
# Get Monaco metadata via celldex
monaco.data = celldex::MonacoImmuneData()
monaco.labels = split(f = monaco.data$label.fine, x = monaco.data$label.ont)
# check if fine-level label to ontology is one to one
print(table(sapply(monaco.labels,function(x){length(unique(x))})))
# make dictionary
monaco_label_fine_to_ont = do.call(
  rbind,
  lapply(names(monaco.labels),function(x){
    CL = gsub(pattern = ":",replacement = "_",x = monaco.labels[[x]][1],fixed = TRUE)
    df = data.frame(from=x,to=CL,is_immune=TRUE)
    if(!df$to%in%hcao_cl_rooted.df$child) df$to = "not hematopoietic"
    return(df)
  })
)

# merge Th1/Th17 cells to Th17 cells based on manual curation of the final HCAO graph
monaco_label_fine_to_ont$to[
  which(monaco_label_fine_to_ont$from=="Th1/Th17 cells")] = "CL_0000899"

# save to file
write.table(monaco_label_fine_to_ont,file = "immune_map_monaco_to_CL.csv",
            quote = FALSE,row.names = FALSE,col.names = TRUE,sep=",")
#
rm(monaco.data)
gc()

```

For Azimuth, we manually build the dictionary by typing the reference page information
in a list.

```{r}
# map Azimuth labels from the reference webpage: https://azimuth.hubmapconsortium.org/references/
azimuth.labels = list(
  "B intermediate" = "CL_0000785",
  "B memory" = "CL_0000787",
  "B naive" = "CL_0000788",
  "Plasmablast" = "CL_0000980",
  "CD4 CTL" = "CL_0000934",
  
  "CD4 Naive" = "CL_0000895",
  "CD4 Proliferating" = "CL_0000896",
  "CD4 TCM" = "CL_0000897",
  "CD4 TEM" = "CL_0001044",
  "Treg" = "CL_0000815",
  
  "CD8 Naive" = "CL_0000900",
  "CD8 Proliferating" = "CL_0000906",
  "CD8 TCM" = "CL_0000909",
  "CD8 TEM" = "CL_0001050",
  "ASDC" = "CL_0000451",
  
  "cDC1" = "CL_0002394",
  "cDC2" = "CL_0002399",
  "pDC" = "CL_0000784",
  "CD14 Mono" = "CL_0002057",
  "CD16 Mono" = "CL_0002396",
  
  "NK" = "CL_0000939",
  "NK Proliferating" = "CL_0000623",
  "NK_CD56bright" = "CL_0000938",
  "Eryth" = "CL_0000764",
  "HSPC" = "CL_0008001",
  
  "ILC" = "CL_0001065",
  "Platelet" = "CL_0000233",
  "dnT" = "CL_0000935",
  "gdT" = "CL_0000798",
  "MAIT" = "CL_0000940"
)

# make dictionary
azimuth_label_fine_to_ont = do.call(
  rbind,
  lapply(names(azimuth.labels),function(x){
    CL = gsub(pattern = ":",replacement = "_",x = azimuth.labels[[x]][1],fixed = TRUE)
    df = data.frame(from=x,to=CL,is_immune=TRUE)
    if(!df$to%in%hcao_cl_rooted.df$child) df$to = "not hematopoietic"
    return(df)
  })
)

# save to file
write.table(azimuth_label_fine_to_ont,file = "immune_map_azimuth_to_CL.csv",
            quote = FALSE,row.names = FALSE,col.names = TRUE,sep=",")
```

## Simplifying HCAO immune graph

We check the overall shape of the HCAO immune graph and how references mapped to it.

```{r}
# check mapped vertices by the 3 references
V(hcao.dag)$azimuth_mapped = V(hcao.dag)$name %in% azimuth_label_fine_to_ont$to
V(hcao.dag)$blueprint_mapped = V(hcao.dag)$name %in% blueprint_label_fine_to_ont$to
V(hcao.dag)$monaco_mapped = V(hcao.dag)$name %in% monaco_label_fine_to_ont$to
V(hcao.dag)$mapped_by = mapply(
  FUN=function(...){
    txt = paste0(c("azimuth","blueprint","monaco")[c(...)],collapse="_")
    if(nchar(txt)<=0) txt = "none"
    return(txt)
    },
  V(hcao.dag)$azimuth_mapped,V(hcao.dag)$blueprint_mapped,V(hcao.dag)$monaco_mapped)
V(hcao.dag)$is_mapped = mapply(
  any,
  V(hcao.dag)$azimuth_mapped,V(hcao.dag)$blueprint_mapped,V(hcao.dag)$monaco_mapped)

# mapping color/shape scheme
mapped_color = c(
  "azimuth" = "#e41a1c",
  "blueprint" = "#377eb8",
  "monaco" = "#FFD700",
  "azimuth_blueprint" = "#984ea3",
  "azimuth_monaco" = "#ff7f00",
  "blueprint_monaco" = "#4daf4a",
  "azimuth_blueprint_monaco" = "black",
  "none" = "grey"
)
mapped_shape = c("TRUE"=16,"FALSE"=1)

# visualization of the mapped DAG
hcao.dag.plot = ggraph(hcao.dag,layout = "kk")+
  geom_edge_link(arrow = arrow(length = unit(2, 'mm'),type = "closed"),color="grey") +
  geom_node_point(mapping = aes(color=mapped_by,shape=is_mapped),size=2) +
  scale_color_manual(values = mapped_color)+
  scale_shape_manual(values = mapped_shape)+
  theme_void()
print(hcao.dag.plot)
```

The full HCAO immune graph is messy and the references have not currently reached
the bottom vertices. As a result, we need to simplify the HCAO immune graph for
practical use of it. Briefly, for each mapped vertex we select the path to the
root that goes through the most mapped vertices and aggregate them as the simplified
HCAO immune graph, with a few exceptions for inclusion/exclusion based on manual
curation of the outcome.

```{r}
### Refine tree to mostly azimuth/blueprint/monaco mapped vertices and resolve multi-parent issues
# also generate the dictionary of cellxgene CL labels to vertices in HCAO DAG

# get all paths to the root
all_paths_to_root = lapply(V(hcao.dag)$name,function(v){
  lapply(all_simple_paths(graph = hcao.dag,from = v,to = "CL_0000988"),as_ids)
})
names(all_paths_to_root) = V(hcao.dag)$name

# preferred vertices to keep in the DAG from manual curation
preferred_vertices = c(
  "CL_0000814", # natural killer T-cell that is not mapped by any reference
  "CL_0000097"  # mast cell that is not mapped by any reference
  )

# Vertices that should be avoided unless they are the only vertices leading to mapped ones
avoided_vertices = c(
  "CL_0000145", # professional antigen presenting cell
  "CL_1001610", # bone marrow hematopoietic cell
  "CL_0000081"  # blood cell that is synonymous to leukocyte
)

# which vertices are mapped by at least one reference
mapped_vertices = V(hcao.dag)$name[V(hcao.dag)$is_mapped]

# calculate the minimal required vertices for the DAG without multiple parent
minimal_edges = unique(do.call(
  rbind,lapply(c(mapped_vertices,preferred_vertices),function(x){
    these_paths = all_paths_to_root[[x]]
    should_avoid = sapply(these_paths,function(xx){any(xx%in%avoided_vertices)})
    if(sum(!should_avoid)>0) these_paths = these_paths[!should_avoid]
    path_lengths = sapply(these_paths,length)
    path_mapped_length = sapply(these_paths,function(xx){
      sum(xx%in%mapped_vertices)
    })
    path_unmapped_length = path_lengths - path_mapped_length
    minimum_length = min(path_lengths)
    max_mapped_length = max(path_mapped_length)
    min_unmapped_length = min(path_unmapped_length)
    ###  adjustable criteria
    candidate_paths = these_paths[path_mapped_length>=max_mapped_length]
    secondary_checked_length = sapply(candidate_paths,length)
    kept_path = candidate_paths[[which.min(secondary_checked_length)]]
    ###
    df = data.frame(
      from = kept_path[-length(kept_path)],
      to = kept_path[-1]
    )
    return(df)
  })
))
minimal_vertices = unique(c(minimal_edges$from,minimal_edges$to))

# add exception of non-immune cells to metadata
hcao_cl_metadata.df = data.frame(name=c("not hematopoietic",minimal_vertices))

# build simplified DAG
hcao.simplified.dag = graph_from_data_frame(
  d = minimal_edges,
  vertices = hcao_cl_metadata.df,
  directed = TRUE)
```

From the simplified HCAO immune graph, we build the dictionary between the original
CELLxGENE annotation and HCAO immune graph.

```{r}
# build cellxgene to HCAO dictionary
cellxgene_CL_to_ont = data.frame(
  from=V(hcao.dag)$name,
  to=sapply(V(hcao.dag)$name,function(x){
    if(x%in%minimal_vertices) return(x)
    these_paths = all_paths_to_root[[x]]
    nearest_vertex_idx = which.min(
      sapply(these_paths,function(xx){
        min(which(xx%in%minimal_vertices))-1
        }))
    y = these_paths[[nearest_vertex_idx]][these_paths[[nearest_vertex_idx]]%in%minimal_vertices]
    return(y[1])
  }),
  is_immune=TRUE)

write.table(cellxgene_CL_to_ont,file = "immune_map_cellxgene_to_CL.csv",
            quote = FALSE,row.names = FALSE,col.names = TRUE,sep=",")

```

## QC of HCAO immune graph

We check the HCAO immune graph visually to make sure it makes the best biological
sense.

```{r}
# count child-parent for each vertex for validation
child_parent_count = do.call(
  rbind,
  lapply(minimal_vertices,
         function(x){
           data.frame(node=x,
                      num_parent=degree(hcao.simplified.dag,x,mode = "out"),
                      num_child =degree(hcao.simplified.dag,x,mode = "in"))
         }))

# check mapped vertices by the 3 references
V(hcao.simplified.dag)$azimuth_mapped = V(hcao.simplified.dag)$name %in% azimuth_label_fine_to_ont$to
V(hcao.simplified.dag)$blueprint_mapped = V(hcao.simplified.dag)$name %in% blueprint_label_fine_to_ont$to
V(hcao.simplified.dag)$monaco_mapped = V(hcao.simplified.dag)$name %in% monaco_label_fine_to_ont$to
V(hcao.simplified.dag)$mapped_by = mapply(
  FUN=function(...){
    txt = paste0(c("azimuth","blueprint","monaco")[c(...)],collapse="_")
    if(nchar(txt)<=0) txt = "none"
    return(txt)
  },
  V(hcao.simplified.dag)$azimuth_mapped,V(hcao.simplified.dag)$blueprint_mapped,V(hcao.simplified.dag)$monaco_mapped)
V(hcao.simplified.dag)$is_mapped = mapply(
  any,
  V(hcao.simplified.dag)$azimuth_mapped,V(hcao.simplified.dag)$blueprint_mapped,V(hcao.simplified.dag)$monaco_mapped)

# add common name label as attribute
V(hcao.simplified.dag)$label = unique_cl_labels[V(hcao.simplified.dag)$name]
V(hcao.simplified.dag)$label[is.na(V(hcao.simplified.dag)$label)] = "non immune"

# visual check
hcao.simp.dag.plot = ggraph(hcao.simplified.dag,layout = "kk")+
  geom_edge_link(arrow = arrow(length = unit(2, 'mm'),type = "closed"),color="grey") +
  geom_node_point(mapping = aes(color=mapped_by,shape=is_mapped),size=2) +
  #geom_node_text(mapping=aes(label=label,color=mapped_by),nudge_y = 0.1)+
  scale_color_manual(values = mapped_color)+
  scale_shape_manual(values = mapped_shape)+
  scale_x_continuous(expand = c(0.1,0.1))+
  theme_void()
print(hcao.simp.dag.plot)
```

## Mapping levels from heuristic immune graph

We reverse the direction of edges for compatibility with the heuristic immune graph.

```{r}
HCAO_graph = reverse_edges(hcao.simplified.dag)
```

We then read the heuristic immune graph and label each node by its level.

```{r}
# load knowledge graph
adj_immune = read.csv("graphs/immune_tree.csv", row.names = 1, check.names = FALSE) |>
  as.matrix()
immune_graph = graph_from_adjacency_matrix(adj_immune, mode = "directed", weighted = TRUE)

# Hierarchy - true (known) hierarchical relationships
# Consensus-only - temporary relationship used for ambiguous or often confused classes (e.g. macrophages and monocytes are often considered a single class but macrophages are not monocytes)
E(immune_graph)$Type = ifelse(E(immune_graph)$weight == 1, "Hierarchy", "Consensus-only")
E(immune_graph)$weight = 1

# Manually curate the graph based on the manuscript
curated_immune_graph = delete_edges(immune_graph,incident(immune_graph, "cytotoxic", mode="out"))

#
immune_graph_to_root = lapply(V(curated_immune_graph)$name,function(nn){
  as_ids(ego(graph = curated_immune_graph,order = -1,nodes = nn,mode = "in")[[1]])
})

#
immune_graph_strata = data.frame(
  label=sapply(immune_graph_to_root,function(x){x[1]}),
  level=paste0("L",sapply(immune_graph_to_root,length)-1))

# Add level to vertex metadata
V(immune_graph)$level = immune_graph_strata$level[match(V(immune_graph)$name,immune_graph_strata$label)]
V(curated_immune_graph)$level = immune_graph_strata$level[match(V(curated_immune_graph)$name,immune_graph_strata$label)]

#
immune_graph_edges = igraph::as_data_frame(immune_graph,what="edges")
immune_graph_vertices = igraph::as_data_frame(immune_graph,what="vertices")

curated_immune_graph_edges = igraph::as_data_frame(curated_immune_graph,what="edges")
curated_immune_graph_vertices = igraph::as_data_frame(curated_immune_graph,what="vertices")

# Immune graph level color scheme
immune_level_color = c(
  L0 = "#001959",
  L1 = "#3B6C55",
  L2 = "#D29243",
  L3 = "#F9CCF9"
)

ggraph(curated_immune_graph,layout = "kk")+
  geom_edge_link(arrow = arrow(length = unit(2, 'mm'),type = "closed"),color="grey") +
  geom_node_point(mapping = aes(color=level),size=3) +
  geom_node_text(mapping=aes(label=name,color=level),nudge_y = 0.1)+
  scale_color_manual(values = immune_level_color)+
  scale_x_continuous(expand = c(0.1,0.1))+
  theme_void()
```

The mapping of heuristic immune graph level is manually done, with the exclusion
of cytotoxic and natural killer T cell.

```{r}
# Manual mapping of immune graph levels to HCAO immune graph
## cytotoxic and nkt were removed based on manual curation
immune_graph_to_HCAO_graph = c(
  "b" = "CL_0000945",
  "b memory" = "CL_0000787",
  "b naive" = "CL_0000788",
  "plasma" = "CL_0000786",
  "ilc" = "CL_0001065",
  "nk" = "CL_0000623",
  "t" = "CL_0000084",
  "t cd4" = "CL_0000624",
  "cd4 naive" = "CL_0000895",
  "cd4 tcm" = "CL_0000904",
  "cd4 tem" = "CL_0000905",
  "cd4 fh em"="CL_0002038",
  "cd4 th1/th17 em"="CL_0000912",
  "cd4 th1 em" = "CL_0000545",
  "cd4 th2 em" = "CL_0000546",
  "cd4 th17 em" = "CL_0000899",
  "t cd8" = "CL_0000625",
  "cd8 naive" = "CL_0000900",
  "cd8 tcm" = "CL_0000907",
  "cd8 tem" = "CL_0000913",
  "tgd" = "CL_0000798",
  "treg" = "CL_0000815",
  "mait" = "CL_0000940",
  "erythrocyte" = "CL_0000764",
  "granulocyte" = "CL_0000094",
  "monocytic" = "CL_0000576",
  "cd14 mono" = "CL_0000860",
  "cd16 mono" = "CL_0000875",
  "dc" = "CL_0000451",
  "cdc" = "CL_0000782",
  "pdc" = "CL_0000784",
  "macrophage" = "CL_0000235",
  "mast" = "CL_0000097",
  "progenitor" = "CL_0008001",
  "non immune" = "not hematopoietic"
)

# Add mapped level to the HCAO immune graph
HCAO_graph_strata = data.frame(
  label=unname(immune_graph_to_HCAO_graph[immune_graph_strata$label]),
  level=immune_graph_strata$level
  )
HCAO_graph_strata = HCAO_graph_strata[!is.na(HCAO_graph_strata$label),]
V(HCAO_graph)$level = HCAO_graph_strata$level[
  match(V(HCAO_graph)$name,HCAO_graph_strata$label)
]
V(HCAO_graph)$n_level = as.numeric(gsub("\\D+", "", V(HCAO_graph)$level))

# Visualize to check
mapped_shape = c("TRUE"=16,"FALSE"=1)
ggraph(HCAO_graph,layout = "kk")+
  geom_edge_link(arrow = arrow(length = unit(2, 'mm'),type = "closed"),color="grey") +
  geom_node_point(mapping = aes(color=level,shape=is_mapped),size=3) +
  #geom_node_text(mapping=aes(label=label,color=level),nudge_y = 0.1)+
  scale_color_manual(values = immune_level_color)+
  scale_shape_manual(values = mapped_shape)+
  scale_x_continuous(expand = c(0.1,0.1))+
  theme_void()
```

Now we fill-in the gap between mapped levels in HCAO graph.

```{r}
# prepare to hierarchically fill-in level for vertices that are not mapped
all_path_to_root = lapply(V(HCAO_graph)$name,function(x){
  lapply(ego(HCAO_graph,order = -1,mode = "in",nodes = x),as_ids)
})
names(all_path_to_root) = V(HCAO_graph)$name

all_children_by_node = lapply(V(HCAO_graph)$name,function(x){
  V(HCAO_graph)$name[sapply(all_path_to_root,function(xx){
    any(sapply(xx,function(xxx){x%in%xxx}))
  })]
})
names(all_children_by_node) = V(HCAO_graph)$name

all_sisters_by_node = lapply(V(HCAO_graph)$name,function(x){
  setdiff(lapply(ego(HCAO_graph,order = 1,mode = "out",nodes = x),as_ids)[[1]],x)
})
names(all_sisters_by_node) = V(HCAO_graph)$name


# fill-in parental vertices (0>1>2)
for(this_level in 0:2){
  level_mapping=V(HCAO_graph)$n_level
  names(level_mapping) = V(HCAO_graph)$name
  vertices_with_level = V(HCAO_graph)$name[which(V(HCAO_graph)$n_level==this_level)]
  parent_until_higher_level = unique(do.call(c,lapply(
    vertices_with_level,
    function(x){
      vv = all_path_to_root[[x]][[1]]
      v2f = vv[which(is.na(level_mapping[vv]))]
      return(v2f)
    }
  )))
  level_mapping[parent_until_higher_level] = this_level
  V(HCAO_graph)$n_level = unname(level_mapping)
}
V(HCAO_graph)$level = sapply(V(HCAO_graph)$n_level,function(x){
  if(is.na(x)) return(NA)
  return(paste0("L",x))
})

# Visualize to check again
ggraph(HCAO_graph,layout = "kk")+
  geom_edge_link(arrow = arrow(length = unit(2, 'mm'),type = "closed"),color="grey") +
  geom_node_point(mapping = aes(color=level,shape=is_mapped),size=3) +
  #geom_node_text(mapping=aes(label=label,color=level),nudge_y = 0.1)+
  scale_color_manual(values = immune_level_color)+
  scale_shape_manual(values = mapped_shape)+
  scale_x_continuous(expand = c(0.1,0.1))+
  theme_void()
```

The remaining unmapped vertices are addressed by their latest mapped ancestors.

```{r}
# now fill-in vertices not mapped and without mapped descendents at level 0-2
vertices_to_fill = V(HCAO_graph)$name[is.na(V(HCAO_graph)$n_level)]
#print(length(vertices_to_fill))
while(length(vertices_to_fill)>0){
  for(this_vertex in vertices_to_fill){
    immediate_parent = all_path_to_root[[this_vertex]][[1]][2]
    # skip if parent not mapped
    if(immediate_parent%in%vertices_to_fill) next
    # prioritize for L0 parent to produce L1 children
    if(level_mapping[immediate_parent]<=0){
      level_mapping[this_vertex] = 1
      next
    }
    # if has mapped sister, use sisters' finest level
    if(length(all_sisters_by_node[[immediate_parent]])>1){
      # ^check if has sister
      # check if any sister is mapped
      sister_level = level_mapping[all_sisters_by_node[[immediate_parent]]]
      sister_level = sister_level[!is.na(sister_level)]
      # use finest level when mapped 
      if(length(sister_level)>0){
        level_mapping[this_vertex] = max(sister_level)
        next
      }
      # fill-in as single child when not mapped
    }
    # +1 if itself is mapped to a reference and +0 if not
    if(V(HCAO_graph)$is_mapped[V(HCAO_graph)$name==this_vertex]){
      level_mapping[this_vertex] = min(level_mapping[immediate_parent]+1,2)
      # limit the extrapolation of level to 2 based on manual curation
    }else{
      level_mapping[this_vertex] = level_mapping[immediate_parent]
    }
  }
  V(HCAO_graph)$n_level = unname(level_mapping)
  vertices_to_fill = V(HCAO_graph)$name[is.na(V(HCAO_graph)$n_level)]
  #print(length(vertices_to_fill))
}

V(HCAO_graph)$level = sapply(V(HCAO_graph)$n_level,function(x){
  if(is.na(x)) return(NA)
  return(paste0("L",x))
})

# Visualize to check again
ggraph(HCAO_graph,layout = "kk")+
  geom_edge_link(arrow = arrow(length = unit(2, 'mm'),type = "closed"),color="grey") +
  geom_node_point(mapping = aes(color=level,shape=is_mapped),size=3) +
  #geom_node_text(mapping=aes(label=label,color=level),nudge_y = 0.1)+
  scale_color_manual(values = immune_level_color)+
  scale_shape_manual(values = mapped_shape)+
  scale_x_continuous(expand = c(0.1,0.1))+
  theme_void()

```

We save the final HCAO graph for use in cellNexus pipeline.

```{r}
HCAO_vertices= igraph::as_data_frame(HCAO_graph,what="vertices")
HCAO_edges= igraph::as_data_frame(HCAO_graph,what="edges")

write.table(HCAO_edges,file = "HCAO_immune_graph_edges.csv",
            quote = TRUE,sep = ",",row.names = FALSE,col.names = TRUE)
write.table(HCAO_vertices[,c("name","is_mapped","mapped_by","label","level")],
            file = "HCAO_immune_graph_vertices.csv",
            quote = TRUE,sep = ",",row.names = FALSE,col.names = TRUE)
```

