---
title: "Ethnicity analysis"
date: "`r Sys.Date()`"
author: "Ning Liu"
format:
  pdf:
    code-fold: true
    toc: true
    css: styles.css
    theme: pulse
    smooth-scroll: true
    code-line-numbers: false
    geometry:
      - left = 20mm
      - right = 20mm
    highlight-style: solarized
    fig-format: png
    fig-dpi: 300
    fig-height: 6
    fig-width: 8
    fig-align: center
    number-sections: true
    number-depth: 4
    fontsize: 14pt
    execute:
      echo: true
      warning: false
      cache: false
    knitr:
      opts_chunk:
        tidy: formatR
        prompt: true
        R.options:
          digits: 4
          width: 80
          scipen: 4
          max.print: 80
editor_options:
  chunk_output_type: inline
---

# Load data

```{r}
library(tidyverse)
library(tidybulk)
library(SummarizedExperiment)
library(SingleCellExperiment)
library(Seurat)
source("../../utils/utils.R")
library(caret)
library(ggpubr)
library(yardstick)
library(Seurat)
library(nnet)
library(ggalluvial)
library(ggsci)
library(patchwork)
library(edgeR)
library(DelayedMatrixStats)
library(cellNexus)
library(igraph)
library(bluster)
```

```{r}
se <- HDF5Array::loadHDF5SummarizedExperiment("pseudobulk_sample")

rd <- as.data.frame(rowData(se)) %>%
  rownames_to_column("gene")

ethnicities <- c(
  "Europeans",
  "EastAsian",
  "SouthAsian",
  "African",
  "HispanicDLatinAmerican"
)

top_markers_all <- map(ethnicities, function(eth) {

  est_col <- paste0("Estimate_", eth)
  se_col  <- paste0("Est.Error_", eth)
  low_col <- paste0("CI.Lower_", eth)
  up_col  <- paste0("CI.Upper_", eth)
  star_col <- paste0("Star_", eth)
  
  rd %>%
    mutate(
      ethnicity = eth,
      estimate = .data[[est_col]],
      std_error = .data[[se_col]],
      ci_lower = .data[[low_col]],
      ci_upper = .data[[up_col]],
      star = if (star_col %in% colnames(.)) .data[[star_col]] else NA,
      dist0_CI = pmax(ci_lower, 0)
    ) %>%
    filter(estimate > 0) %>%
    arrange(desc(dist0_CI), desc(estimate)) %>%
    slice_head(n = 200) %>%
    dplyr::select(
      ethnicity, gene,
      estimate, std_error,
      ci_lower, ci_upper,
      dist0_CI, star
    )
}) %>%
  bind_rows()

selected_marker_genes <- unique(top_markers_all$gene)

sce_all <- SingleCellExperiment(assay = list("counts" = assay(se, 1),
                                             "gene_presence" = assay(se, 2),
                                             "counts_scaled" = assay(se, 3),
                                             "counts_adjusted_ethnicity" = assay(se, 4)),
                                colData = colData(se),
                                rowData = rowData(se))

rowData(sce_all)$is_ethnicity_marker <- rownames(sce_all) %in% selected_marker_genes

sce <- sce_all[selected_marker_genes,]
```


```{r}
metadata <- get_metadata()

check_eth <- metadata |>
  dplyr::select(dataset_id, self_reported_ethnicity) %>%
  as_tibble()

check_eth2 <- metadata |>
  dplyr::select(dataset_id, sample_id, self_reported_ethnicity, self_reported_ethnicity_ontology_term_id) %>%
  as_tibble()

chosen_samples <- check_eth2 %>%
  filter(sample_id %in% sce$sample_id) %>%
  dplyr::select(c("sample_id","self_reported_ethnicity")) %>%
  unique()

taxonomy_lookup <- tribble(
  ~self_reported_ethnicity, ~L1,      ~L2,            ~L3,                         ~Leaf,
  
  "unknown",                "World",  "Unknown",      "Unknown",                   "unknown",
  
  "African",                "World",  "African",      "Unspecified",               "African",
  "African American",       "World",  "African",      "Diaspora",                  "African American",
  "African American or Afro-Caribbean",
  "World",  "African",      "Diaspora",                  "Afro-Caribbean",
  
  "European",               "World",  "European",     "Unspecified",               "European",
  "European American",      "World",  "European",     "Diaspora",                  "European American",
  
  "Finnish",                "World",  "Nordic",       "Unspecified",               "Finnish",
  
  "British",                "World",  "CelticEnglish","Unspecified",               "British",
  "Irish",                  "World",  "CelticEnglish","Unspecified",               "Irish",
  
  "Jewish Israeli",         "World",  "Jewish",       "Israel",                    "Israel",
  
  
  "East Asian",             "World",  "EastAsian",    "Unspecified",               "East Asian",
  "Chinese",                "World",  "EastAsian",    "Chinese",                   "Chinese",
  "Han Chinese",            "World",  "EastAsian",    "Chinese",                   "Han Chinese",
  "Singaporean Chinese",    "World",  "EastAsian",    "Chinese diaspora",          "Singaporean Chinese",
  "Japanese",               "World",  "EastAsian",    "Japan",                     "Japanese",
  "Korean",                 "World",  "EastAsian",    "Korea",                     "Korean",
  
  "South Asian",            "World",  "SouthAsian",   "Unspecified",               "South Asian",
  "Bangladeshi",            "World",  "SouthAsian",   "Bangladesh",                "Bangladeshi",
  "Singaporean Indian",     "World",  "SouthAsian",   "Indian diaspora",           "Singaporean Indian",
  
  "South East Asian",       "World",  "EastAsian",    "Malay/SEAsia",              "South East Asian",
  "Singaporean Malay",      "World",  "EastAsian",    "Malay/SEAsia",              "Singaporean Malay",
  
  "Hispanic or Latin American",
  "World",  "Hispanic",     "Spanish/Latin America",     "Hispanic or Latin American",
  
  "Iraqi",                  "World",  "Muslim/MENA",  "Arabian Peninsula/MENA",    "Iraqi",
  "Greater Middle Eastern  (Middle Eastern or North African or Persian)",
  "World",  "Muslim/MENA",  "MENA/Persian (broad)",      "Greater Middle Eastern",
  
  "Oceanian",               "World",  "Oceanian",     "Oceania",                   "Oceanian",
  "Pacific Islander",       "World",  "Oceanian",     "Pacific Islander",          "Pacific Islander",
  
  "Asian",                  "World",  "Asian",        "Unspecified",               "Asian",
  "American",               "World",  "Americas",     "North America",             "American",
  
  "European,Asian",         "World",  "Mixed",        "Europe + Asia",             "European,Asian",
  "Native American,Hispanic or Latin American",
  "World",  "Mixed",        "Native + Hispanic",         "Native American + Hispanic"
)

chosen_samples <- chosen_samples %>%
  left_join(taxonomy_lookup, by = "self_reported_ethnicity") %>%
  mutate(
    L1   = coalesce(L1, "World"),
    L2   = coalesce(L2, "Other/Unmapped"),
    L3   = coalesce(L3, "Other/Unmapped"),
    Leaf = coalesce(Leaf, self_reported_ethnicity)
  )

```


```{r}

unknown_samples <- chosen_samples %>%
  filter(!L2 %in% c("African","EastAsian","European","Hispanic","SouthAsian"))

chosen_samples <- chosen_samples %>%
  filter(L2 %in% c("African","EastAsian","European","Hispanic","SouthAsian"))

colnames(sce) <- sce$sample_id
colnames(sce_all) <- sce_all$sample_id

sce_query <- sce[,sce$sample_id %in% unknown_samples$sample_id]
sce_query_all <- sce_all[,sce_all$sample_id %in% unknown_samples$sample_id]
sce_query <- sce_query_all

sce <- sce[,sce$sample_id %in% chosen_samples$sample_id]
sce_all <- sce_all[,sce_all$sample_id %in% chosen_samples$sample_id]

sce <- sce[,chosen_samples$sample_id]
sce_all <- sce_all[,chosen_samples$sample_id]

all(colnames(sce) == chosen_samples$sample_id)
all(colnames(sce_all) == chosen_samples$sample_id)

sce$ethnicity_groups <- chosen_samples$L2
sce_all$ethnicity_groups <- chosen_samples$L2
```


```{r}
eth_tissue_heatmap1 <- table(colData(se)[,c("ethnicity_groups","tissue_groups")]) %>%
  as.matrix() %>%
  log1p() %>%
  scale() %>%
  t() %>%
  scale() %>%
  t() %>%
  pheatmap::pheatmap(border_color = "white", cellwidth = 15, cellheight = 15, color = scico::scico(256, direction = -1, palette = "roma"))

eth_tissue_heatmap2 <- table(colData(sce)[,c("ethnicity_groups","tissue_groups")]) %>%
  as.matrix() %>%
  log1p() %>%
  scale() %>%
  t() %>%
  scale() %>%
  t() %>%
  pheatmap::pheatmap(border_color = "white", cellwidth = 15, cellheight = 15, color = scico::scico(256, direction = -1, palette = "roma"))
```

```{r}
#| fig-width: 10
#| fig-height: 10
ggplotify::as.ggplot(eth_tissue_heatmap1) + 
  ggplotify::as.ggplot(eth_tissue_heatmap2) +
  plot_layout(ncol = 1, widths = c(1,1)) +
  plot_annotation(tag_levels = "A")

ggsave("../output/ethnicity_tissue_distribution_heatmap.png", width = 10, height = 10, bg = "white")
```


```{r}
pca <- prcomp(t(assay(sce, "counts_adjusted_ethnicity")), rank. = 30, scale. = TRUE)
umap <- uwot::tumap(pca$x[,1:10], n_components = 2)
colnames(umap) <- paste0("UMAP",seq(2))


reducedDim(sce, "PCA") <- pca$x
reducedDim(sce, "UMAP") <- umap
```

```{r}
npg_cols <- c(ggsci::pal_npg("nrc")(5),"gray")
my_colors <- setNames(npg_cols, 
  c("European",
    "EastAsian",
    "SouthAsian",
    "African",
    "Hispanic",
    "LowConfidenceLabel")
)
```

# check umap with ehtnicity labels

UMAP on adjusted count

```{r}
#| fig-width: 6
#| fig-height: 7
standR::plotDR(sce, dimred = "UMAP", 
               color = ethnicity_groups, shape = 16,
               size = 1) +
  scale_color_manual(values = my_colors) +
  theme_void() +
  labs(x = "UMAP1", y = "UMAP2") +
  guides(color=guide_legend(override.aes = list(size = 5), nrow = 3)) +
  theme(legend.title=element_text(size=17), 
    legend.text=element_text(size=14), legend.position = "bottom",
    panel.border = element_rect(color = "black", fill = NA, size = 2),
    plot.margin = unit(c(10,30,10,20), "pt"))
```


```{r}
#| fig-width: 8
#| fig-height: 8
p_post_batch_correct <- standR::plotDR(sce, dimred = "UMAP", 
               color = ethnicity_groups, shape = 16,
               size = 1) +
  scale_color_manual(values = my_colors) +
  theme_void() +
  labs(x = "UMAP1", y = "UMAP2") +
  guides(color=guide_legend(override.aes = list(size = 5), nrow = 3)) +
  theme(legend.title=element_text(size=17), 
    legend.text=element_text(size=14), legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA, size = 2),
    plot.margin = unit(c(10,10,10,10), "pt"))

p_post_batch_correct

ggsave(p_post_batch_correct, file = "../output/umap_original_adjusted.pdf", width = 8, height = 7)
```

# check umap with uncorrected data

```{r}
sce_ua <- sce

sce_ua <- scater::runPCA(sce_ua, exprs_values = "counts")
sce_ua <- scater::runUMAP(sce_ua, dimred = "PCA")
```

```{r}
#| fig-width: 9
#| fig-height: 9
p_before_batch_correct <- standR::plotDR(sce_ua, dimred = "UMAP", 
               color = ethnicity_groups, shape = 16,
               size = 1) +
  scale_color_manual(values = my_colors) +
  theme_void() +
  labs(x = "UMAP1", y = "UMAP2") +
  guides(color=guide_legend(override.aes = list(size = 5), nrow = 4)) +
  theme(legend.title=element_text(size=17), 
    legend.text=element_text(size=14),
    legend.position = "none",
    panel.border = element_rect(color = "black", fill = NA, size = 2),
    plot.margin = unit(c(10,10,10,10), "pt"))

p_before_batch_correct

ggsave(p_before_batch_correct, file = "../output/umap_original_unadjusted.pdf", width = 8, height = 7)
```


# assign a confidence score to each sample

## local label agreement & final score

The previous version penalised every sample in a high-entropy cluster equally. In practice that is too blunt: if a branch point and a clean branch are merged into the same cluster, the whole branch can be pushed into the ambiguous class. To avoid that failure mode, we keep the cluster entropy only as a diagnostic and compute the final confidence score from **cell-level neighbourhood structure**.

For **each sample** $i$:

1.  Find the exact $k_{\mathrm{local}}$ nearest neighbours of $i$ in PCA space.

2.  Compute the neighbourhood label distribution $p_i(\ell)$.

3.  Define:

$$
   \alpha_i = p_i(L_i)
$$

as the local purity for the sample's assigned label, and

$$
   m_i = \alpha_i - \max_{\ell \neq L_i} p_i(\ell)
$$

as the local margin over the strongest competing label.

4.  Use the final confidence score

$$
   \mathrm{score}_i = \alpha_i \times \frac{1 + m_i}{2}.
$$

This keeps coherent branches high-confidence when their local neighbourhood is pure, while only the branch junction and other locally mixed regions become ambiguous.

```{r}
set.seed(10)

pca_data <- pca$x
df <- colData(sce) %>% as.data.frame()
k_local <- 15
knn_res <- BiocNeighbors::findKNN(pca_data, k = k_local)
nn_index <- knn_res$index
label_levels <- sort(unique(as.character(df$ethnicity_groups)))

calc_local_label_score <- function(label_i, neighbor_labels, label_levels) {
  label_i <- as.character(label_i)
  prob_tbl <- table(factor(neighbor_labels, levels = label_levels))
  probs <- as.numeric(prob_tbl) / sum(prob_tbl)
  names(probs) <- label_levels

  alpha_i <- unname(probs[label_i])
  other_probs <- probs[names(probs) != label_i]
  runner_up <- if (length(other_probs)) max(other_probs) else 0
  margin_i <- alpha_i - runner_up

  score_i <- alpha_i * ((margin_i + 1) / 2)

  tibble(
    alpha = alpha_i,
    runner_up = runner_up,
    margin = margin_i,
    score = score_i
  )
}

local_stats <- vector("list", nrow(pca_data))

for (i in seq_len(nrow(pca_data))) {
  nn_ids <- nn_index[i, ]
  neighbor_labels <- as.character(df$ethnicity_groups[nn_ids])
  local_stats[[i]] <- calc_local_label_score(
    label_i = df$ethnicity_groups[i],
    neighbor_labels = neighbor_labels,
    label_levels = label_levels
  )
}

local_stats <- bind_rows(local_stats)

df <- bind_cols(df, local_stats)
colData(sce)$alpha <- df$alpha
colData(sce)$runner_up <- df$runner_up
colData(sce)$margin <- df$margin
colData(sce)$local_entropy <- df$local_entropy
colData(sce)$local_entropy_norm <- df$local_entropy_norm
colData(sce)$score <- df$score
```

```{r}
#| fig-width: 8
#| fig-height: 8
standR::plotDR(sce, dimred = "UMAP", 
               color = score, shape = 16,
               size = 1.5) +
  scale_color_gradient(low = "red", high = "gray") +
  theme_ning() +
  labs(x = "UMAP1", y = "UMAP2") +
  theme(legend.key.width=unit(1,"cm"))
```

```{r}
p_scoring_umap <- standR::plotDR(sce, dimred = "UMAP", 
               color = score, shape = 16,
               size = 1.5) +
  scale_color_gradient(low = "red", high = "gray") +
  theme_void() +
  labs(x = "UMAP1", y = "UMAP2") +
  theme(legend.key.width=unit(1,"cm"),
        legend.position = "none",
        panel.border = element_rect(color = "black", fill = NA, size = 2),
        plot.margin = unit(c(10,10,10,10), "pt"))

p_scoring_umap

ggsave(p_scoring_umap, file = "../output/umap_score.pdf", width = 8, height = 7)
```



## outputing relabels.

```{r}
confidence_threshold <- 0.55

sce$ethnicity_relabel <- colData(sce) %>%
  as.data.frame() %>%
  mutate(ethnicity_relabel = ifelse(score >= confidence_threshold, as.character(ethnicity_groups), "LowConfidenceLabel")) %>%
  .$ethnicity_relabel

colData(sce) %>% 
  as.data.frame() %>% 
  dplyr::select(c("sample_id", "ethnicity_relabel")) %>% 
  mutate(Filter = ifelse(ethnicity_relabel == "LowConfidenceLabel", TRUE, FALSE)) %>%
  write_csv("../output/sample_filtering_based_on_ethnicity_clearness.csv")
```




```{r}
#| fig-width: 9
#| fig-height: 9
standR::plotDR(sce, dimred = "UMAP", 
               color = ethnicity_relabel, shape = 16,
               size = .8) +
  scale_color_manual(values = my_colors, name = "ethnicity_groups") +
  theme_ning() +
  labs(x = "UMAP1", y = "UMAP2") +
  guides(color=guide_legend(override.aes = list(size = 5), nrow = 4)) +
  theme(legend.title=element_text(size=17), 
    legend.text=element_text(size=14))
```

```{r}
sce$UMAP1 <- reducedDim(sce, "UMAP")[,"UMAP1"]
sce$UMAP2 <- reducedDim(sce, "UMAP")[,"UMAP2"]


colData(sce) %>% as.data.frame() %>% filter(ethnicity_relabel == "EastAsian") %>% filter(UMAP1 < 2) %>% rownames()
colData(sce) %>% as.data.frame() %>% filter(ethnicity_relabel == "European") %>% filter(UMAP1 > 7) %>% rownames()
```

```{r}
samples2check <- c(colData(sce) %>% as.data.frame() %>% filter(ethnicity_relabel == "EastAsian") %>% filter(UMAP1 < 2) %>% rownames() %>% head(n=2), 
                   colData(sce) %>% as.data.frame() %>% filter(ethnicity_relabel == "European") %>% filter(UMAP1 > 7) %>% rownames() %>% head(n = 2))
```

```{r}
samples2check_df <- reducedDim(sce, "UMAP")[samples2check,] %>%
  cbind(colData(sce) %>% 
          as.data.frame() %>% 
          .[samples2check,]) %>%
  as.data.frame() %>%
  dplyr::select(c("sample_id","UMAP1","UMAP2","ethnicity_relabel","ethnicity_groups"))
```

6 suspicious samples are selected.

```{r}
#| fig-width: 9
#| fig-height: 9
p_sus1 <- standR::plotDR(sce, dimred = "UMAP", 
               color = ethnicity_relabel, shape = 16,
               size = .8) +
  scale_color_manual(values = my_colors, name = "ethnicity_groups") +
  geom_point(data = samples2check_df, aes(x = UMAP1, y = UMAP2), color = "black", shape = 1, size = 2) +
  ggrepel::geom_label_repel(data = samples2check_df, aes(x = UMAP1, y = UMAP2, label = ethnicity_relabel)) + 
  theme_ning() +
  labs(x = "UMAP1", y = "UMAP2") +
  guides(color=guide_legend(override.aes = list(size = 5), nrow = 2)) +
  theme(legend.title=element_text(size=17), 
    legend.text=element_text(size=14))

p_sus1

ggsave("../output/sus_case_umap.png", width = 11, height = 11, bg = "white")
```

```{r}
set.seed(10)
g <- makeKNNGraph(pca_data, k = 10, directed = FALSE)
V(g)$name <- rownames(pca_data)

dfl <- list()

for(i in samples2check){
  v_index <- which(V(g)$name == i)
  node_neighbors <- neighbors(g, v_index)
  neighbor_names <- V(g)$name[node_neighbors]
  df1 <- data.frame(target = i,
            sampleID = c(i,neighbor_names))
  
  dfl[[i]] <- df1
}
```


```{r}
plot_list <- lapply(dfl, function(x){
  as.data.frame(colData(sce[,x$sampleID])) %>%
    rownames_to_column("sampleID") %>%
    left_join(x) %>%
    ggplot(aes(UMAP1, UMAP2, color = ethnicity_groups)) +
    geom_point(data = as.data.frame(colData(sce)), aes(UMAP1, UMAP2), color = "gray", size = 0.1) +
    geom_point(size = 2.5) +
    geom_point(data = samples2check_df %>% 
                 filter(sample_id == unique(x$target)), 
               aes(UMAP1, UMAP2), color = "black", shape = 1, size = 2.5) +
    scale_color_manual(values = my_colors, name = "ethnicity_groups") +
    theme_ning() +
    labs(x = "UMAP1", y = "UMAP2") +
    guides(color=guide_legend(override.aes = list(size = 5), nrow = 1)) +
    theme(legend.title=element_text(size=12), 
      legend.text=element_text(size=11)) +
    theme(legend.position = "none")
})
```

We can see the neighbors of each sample are of the same colors, indicating the same ethnicity, so they should high confidence, consistent with scoring algorithm. The misplacement on the UMAP space is just because it's not a linear transformation.

Red circles are the 6 suspicious samples.

```{r}
#| fig-width: 15
#| fig-height: 11
p_sus2 <- ggpubr::ggarrange(plotlist = plot_list)
p_sus2 <- annotate_figure(p_sus2, top = text_grob("Visualise neighbors of specific samples with the kNN-Graph", 
               color = "black", face = "bold", size = 15))

p_sus2

ggsave("../output/sus_case_umap_knn_highlight.png", width = 15, height = 11, bg = "white")
```


```{r}
#| fig-width: 8
#| fig-height: 13
p_sus1 + p_sus2 + plot_layout(ncol = 1, heights = c(1,1)) + plot_annotation(tag_levels = "A")

ggsave("../output/suspicous_sample_qc_umaps.png", width = 7, height = 13, bg = "white")
```


Here we plot without the low confidence label samples.

```{r}
#| fig-width: 9
#| fig-height: 9
p_onlyhighscore <- standR::plotDR(sce[, sce$ethnicity_relabel!="LowConfidenceLabel"], dimred = "UMAP", 
               color = ethnicity_relabel, shape = 16,
               size = 1.3) +
  scale_color_manual(values = my_colors, name = "ethnicity_groups") +
  theme_ning() +
  labs(x = "UMAP1", y = "UMAP2") +
  guides(color=guide_legend(override.aes = list(size = 8, shape = 15), nrow = 4)) +
  theme(legend.title=element_text(size=20), 
    legend.text=element_text(size=18),
    axis.text = element_text(size=18),
    axis.title = element_text(size=22)) +
  theme_void() +
  theme(legend.position = "none",
        panel.border = element_rect(color = "black", fill = NA, size = 2),
        plot.margin = unit(c(10,10,10,10), "pt"))

p_onlyhighscore

ggsave(p_onlyhighscore, file = "../output/UMAP_highConfidence.pdf", width = 8, height = 7)
```

```{r}
transfer_cols <- c(
  "ethnicity_groups",
  "clusters",
  "entropy",
  "entropy_study",
  "alpha",
  "runner_up",
  "margin",
  "local_entropy",
  "local_entropy_norm",
  "score",
  "ethnicity_relabel"
)

for (col_nm in transfer_cols) {
  colData(sce_all)[[col_nm]] <- colData(sce)[[col_nm]][match(colnames(sce_all), colnames(sce))]
}

add_training_assays <- function(sce_obj) {
  raw_counts <- as.matrix(assay(sce_obj, "counts"))

  assay(sce_obj, "logcounts") <- edgeR::cpm(raw_counts, log = TRUE)

  sce_tmm <- sce_obj %>%
    tidybulk::scale_abundance(method = "TMM", suffix = "_tmm")

  if (!"counts_tmm" %in% assayNames(sce_tmm)) {
    stop("counts_tmm assay was not created by tidybulk::scale_abundance().")
  }

  counts_tmm <- as.matrix(assay(sce_tmm, "counts_tmm"))

  assay(sce_obj, "counts_tmm") <- counts_tmm
  assay(sce_obj, "logcounts_tmm") <- log1p(counts_tmm)
  assay(sce_obj, "cpm_tmm_logcounts") <- edgeR::cpm(counts_tmm, log = TRUE)

  sce_obj
}

sce_all <- add_training_assays(sce_all)
sce_query_all <- add_training_assays(sce_query_all)
```


# output h5ad

```{r}
zellkonverter::writeH5AD(sce_all, file = "../data/sce_relabel.h5ad", compression = "gzip")

sce <- zellkonverter::readH5AD("../data/sce_relabel.h5ad")
```

```{r}
sce_ref <- sce[,sce$ethnicity_relabel!="LowConfidenceLabel"]
```

```{r}
sce_l <- sce[, sce$ethnicity_relabel == "LowConfidenceLabel"]
```

```{r}
sce_query
```

```{r}
intersect_cols <- intersect(colnames(colData(sce_l)),
                            colnames(colData(sce_query)))
```

```{r}
cold_merge_query <- colData(sce_l)[,intersect_cols] %>%
  rbind(colData(sce_query)[,intersect_cols]) %>%
  as.data.frame()
  
```

```{r}
sce_query <- SingleCellExperiment(assay = list("counts" = 
                                               cbind(as.data.frame(assay(sce_l,"counts")),
                                                     as.data.frame(assay(sce_query,"counts"))),
                                             "counts_adjusted_ethnicity" = 
                                               cbind(as.data.frame(assay(sce_l,"counts_adjusted_ethnicity")),
                                                     as.data.frame(assay(sce_query,"counts_adjusted_ethnicity"))),
                                             "counts_scaled" = 
                                               cbind(as.data.frame(assay(sce_l,"counts_scaled")),
                                                     as.data.frame(assay(sce_query,"counts_scaled"))),
                                             "gene_presence" = 
                                               cbind(as.data.frame(assay(sce_l,"gene_presence")),
                                                     as.data.frame(assay(sce_query,"gene_presence")))),
                                colData = cold_merge_query)

seu_query <- CreateSeuratObject(counts = as.matrix(assay(sce_query, "counts")),
                              meta.data = as.data.frame(colData(sce_query)))

seu_query@assays$RNA$data <- as.matrix(assay(sce_query, "counts_adjusted_ethnicity"))
```

```{r}
set.seed(42)

pca_q <- prcomp(t(assay(sce_query, "counts_adjusted_ethnicity")), rank. = 30, scale. = TRUE)
umap_q <- uwot::tumap(pca_q$x[,1:10], n_components = 2)
colnames(umap_q) <- paste0("UMAP",seq(2))

reducedDim(sce_query, "PCA") <- pca_q$x
reducedDim(sce_query, "UMAP") <- umap_q
```

```{r}
p_query_umap <- standR::plotDR(sce_query, dimred = "UMAP", 
               color = "black", shape = 16,
               size = 1) +
  theme_void() +
  theme(panel.border = element_rect(color = "black", fill = NA, size = 2),
        plot.margin = unit(c(10,10,10,10), "pt"))

p_query_umap

ggsave(p_query_umap, file = "../output/UMAP_query.pdf", width = 8, height = 7)
```


```{r}
zellkonverter::writeH5AD(sce_query_all, file = "../data/sce_query.h5ad", compression = "gzip")

sce_query <- zellkonverter::readH5AD(file = "../data/sce_query.h5ad")
```

Now need to run the Classifier_benchmark.ipynb file

# Benchmarking prediction methods 

## Seurat label transfer

```{r}
sce_ref <- sce[, sce$ethnicity_relabel != "LowConfidenceLabel"]
assay(sce_ref, "logcounts") <- as.matrix(assay(sce_ref, "logcounts"))
assay(sce_ref, "counts") <- as.matrix(assay(sce_ref, "counts"))
seu_ref <- as.Seurat(sce_ref)
levels_order <- unique(sce_ref$ethnicity_relabel)
clean_levels <- make.names(levels_order, unique = TRUE)
clean_to_ori_name <- levels_order
names(clean_to_ori_name) <- clean_levels
seu_ref$ethnicity_clean <- factor(seu_ref$ethnicity_relabel, levels = levels_order, labels = clean_levels)
split_seed <- 52
n_folds <- 5
benchmark_recipe_names <- c(
  "LogCPMScale",
  "LogCPMPCA",
  "Log1pTMMScale",
  "Log1pTMMPCA",
  "LogCPM_TMMScale",
  "LogCPM_TMMPCA"
)
selected_models <- c(
  as.vector(t(outer(c("HistGBDT", "LinearSVM", "LogReg", "MLP", "RandomForest"), benchmark_recipe_names, paste, sep = "__"))),
  "RandomForestRank__LogCPMRank"
)
```

```{r}
build_group_holdout_folds <- function(meta_df, group_col, k = 5, seed = 52, protected_groups = character()) {
  set.seed(seed)

  if (!group_col %in% colnames(meta_df)) {
    stop(group_col, " is required in meta_df.")
  }

  df <- meta_df %>%
    filter(!is.na(.data[[group_col]]), !is.na(ethnicity)) %>%
    transmute(
      holdout_group = as.character(.data[[group_col]]),
      ethnicity = as.character(ethnicity)
    )

  all_ethnicities <- sort(unique(df$ethnicity))

  group_stats <- df %>%
    dplyr::count(holdout_group, ethnicity, name = "n") %>%
    pivot_wider(names_from = ethnicity, values_from = n, values_fill = 0)

  group_sizes <- df %>%
    dplyr::count(holdout_group, name = "n_samples")

  group_stats <- group_sizes %>%
    left_join(group_stats, by = "holdout_group")

  class_group_counts <- colSums(as.matrix(group_stats[, all_ethnicities, drop = FALSE]) > 0)

  coverage_list <- setNames(
    lapply(seq_len(nrow(group_stats)), function(i) {
      all_ethnicities[as.numeric(group_stats[i, all_ethnicities, drop = TRUE]) > 0]
    }),
    group_stats$holdout_group
  )

  group_names <- group_stats$holdout_group
  protected_lookup <- tolower(protected_groups)
  candidate_group_names <- group_names[!tolower(group_names) %in% protected_lookup]
  total_samples <- sum(group_stats$n_samples)
  target_test_n <- ceiling(total_samples / k)
  usage_counts <- setNames(integer(length(group_names)), group_names)
  folds <- vector("list", k)
  heldout_groups <- vector("list", k)

  for (i in seq_len(k)) {
    selected <- character()
    covered <- character()
    current_n <- 0

    repeat {
      remaining_ethnicities <- setdiff(all_ethnicities, covered)
      candidates <- setdiff(candidate_group_names, selected)

      if (!length(remaining_ethnicities) && current_n >= target_test_n) {
        break
      }
      if (!length(candidates)) {
        break
      }

      candidate_tbl <- tibble(holdout_group = candidates) %>%
        mutate(
          gain = map_int(holdout_group, ~ sum(remaining_ethnicities %in% coverage_list[[.x]])),
          rarity_gain = map_dbl(holdout_group, function(grp) {
            gained <- intersect(remaining_ethnicities, coverage_list[[grp]])
            if (!length(gained)) {
              return(0)
            }
            sum(1 / class_group_counts[gained])
          }),
          keeps_train_cover = map_lgl(holdout_group, function(grp) {
            train_groups <- setdiff(group_names, c(selected, grp))
            train_cover <- unique(unlist(coverage_list[train_groups]))
            all(all_ethnicities %in% train_cover)
          }),
          usage = usage_counts[holdout_group],
          n_samples = group_stats$n_samples[match(holdout_group, group_names)],
          size_gap = abs((current_n + n_samples) - target_test_n),
          tie_break = runif(n())
        ) %>%
        filter(keeps_train_cover)

      if (!nrow(candidate_tbl)) {
        break
      }

      if (length(remaining_ethnicities)) {
        candidate_tbl <- candidate_tbl %>% arrange(desc(gain), desc(rarity_gain), usage, size_gap, tie_break)
      } else {
        candidate_tbl <- candidate_tbl %>% arrange(usage, size_gap, desc(rarity_gain), tie_break)
      }

      chosen <- candidate_tbl$holdout_group[1]
      selected <- c(selected, chosen)
      covered <- union(covered, coverage_list[[chosen]])
      current_n <- current_n + group_stats$n_samples[match(chosen, group_names)]
      usage_counts[chosen] <- usage_counts[chosen] + 1L
    }

    train_groups <- setdiff(group_names, selected)
    train_cover <- unique(unlist(coverage_list[train_groups]))

    folds[[i]] <- which(as.character(meta_df[[group_col]]) %in% selected)
    heldout_groups[[i]] <- selected
  }

  names(folds) <- paste0("Fold", seq_len(k))
  attr(folds, "heldout_groups") <- heldout_groups
  folds
}

make_split_plan <- function(seu_obj, group_col, k = 5, seed = 52) {
  group_values <- as.character(seu_obj@meta.data[[group_col]])
  fold_meta <- tibble(
    sample_id = colnames(seu_obj),
    ethnicity = as.character(seu_obj$ethnicity_clean),
    group_value = group_values
  )
  colnames(fold_meta)[colnames(fold_meta) == "group_value"] <- group_col

  protected_groups <- if (identical(group_col, "tissue_groups")) "blood" else character()

  folds <- build_group_holdout_folds(
    fold_meta,
    group_col = group_col,
    k = k,
    seed = seed,
    protected_groups = protected_groups
  )

  for (i in seq_along(folds)) {
    test_i <- folds[[i]]
    train_i <- setdiff(seq_len(ncol(seu_obj)), test_i)
    stopifnot(length(intersect(unique(group_values[test_i]), unique(group_values[train_i]))) == 0)
    stopifnot(all(clean_levels %in% unique(as.character(seu_obj$ethnicity_clean[test_i]))))
    stopifnot(all(clean_levels %in% unique(as.character(seu_obj$ethnicity_clean[train_i]))))
  }

  list(
    group_col = group_col,
    folds = folds,
    heldout_groups = attr(folds, "heldout_groups"),
    group_values = group_values
  )
}

split_plan_list <- list(
  dataset_id = make_split_plan(seu_ref, group_col = "dataset_id", k = n_folds, seed = split_seed),
  tissue_groups = make_split_plan(seu_ref, group_col = "tissue_groups", k = n_folds, seed = split_seed)
)

folds <- split_plan_list$dataset_id$folds
heldout_dataset_list <- split_plan_list$dataset_id$heldout_groups
groups <- split_plan_list$dataset_id$group_values
```

```{r}
cv_split_group_df <- imap_dfr(
  split_plan_list,
  function(plan, split_scheme) {
    map2_dfr(
      plan$folds,
      seq_along(plan$folds),
      ~ {
        test_i <- .x
        train_i <- setdiff(seq_len(ncol(seu_ref)), test_i)

        bind_rows(
          tibble(
            split_scheme = split_scheme,
            fold = .y,
            split = "Train",
            holdout_group = plan$group_values[train_i]
          ),
          tibble(
            split_scheme = split_scheme,
            fold = .y,
            split = "Test",
            holdout_group = plan$group_values[test_i]
          )
        )
      }
    )
  }
) %>%
  dplyr::count(split_scheme, fold, split, holdout_group, name = "n_samples") %>%
  group_by(split_scheme, fold, split) %>%
  mutate(prop_samples = n_samples / sum(n_samples)) %>%
  ungroup()

cv_split_group_df

holdout_group_order <- cv_split_group_df %>%
  group_by(split_scheme, holdout_group) %>%
  summarise(n_total = sum(n_samples), .groups = "drop") %>%
  arrange(split_scheme, desc(n_total))

cv_split_group_plot_df <- cv_split_group_df %>%
  left_join(holdout_group_order, by = c("split_scheme", "holdout_group")) %>%
  group_by(split_scheme) %>%
  mutate(holdout_group = factor(holdout_group, levels = rev(unique(holdout_group[order(n_total)])))) %>%
  ungroup()

ggplot(cv_split_group_plot_df %>% filter(split_scheme == "dataset_id"), aes(x = factor(fold), y = holdout_group, fill = prop_samples)) +
  geom_tile(color = "white") +
  facet_wrap(~ split, scales = "free_y", nrow = 1) +
  scale_fill_viridis_c(labels = scales::percent_format(accuracy = 1)) +
  theme_classic() +
  theme(axis.text.y = element_blank(),
        axis.ticks.y = element_blank()) +
  labs(
    x = "Cross-validation fold",
    y = "Studies",
    fill = "Within-split proportion",
    title = "Train/test group composition across held-out folds"
  )

ggsave("../output/study_held_out_distribution_check.pdf", width = 7, height = 7, bg = "white")

cv_split_ethnicity_df <- imap_dfr(
  split_plan_list,
  function(plan, split_scheme) {
    map2_dfr(
      plan$folds,
      seq_along(plan$folds),
      ~ {
        test_i <- .x
        train_i <- setdiff(seq_len(ncol(seu_ref)), test_i)

        bind_rows(
          tibble(
            split_scheme = split_scheme,
            fold = .y,
            split = "Train",
            ethnicity = as.character(seu_ref$ethnicity_relabel[train_i])
          ),
          tibble(
            split_scheme = split_scheme,
            fold = .y,
            split = "Test",
            ethnicity = as.character(seu_ref$ethnicity_relabel[test_i])
          )
        )
      }
    )
  }
) %>%
  dplyr::count(split_scheme, fold, split, ethnicity, name = "n_samples") %>%
  group_by(split_scheme, fold, split) %>%
  mutate(prop_samples = n_samples / sum(n_samples)) %>%
  ungroup()

ggplot(cv_split_ethnicity_df %>% filter(split_scheme == "dataset_id"), aes(x = factor(fold), y = prop_samples, fill = ethnicity)) +
  geom_col(color = "white") +
  facet_wrap( ~ split) +
  scale_fill_manual(values = my_colors[names(my_colors) %in% unique(cv_split_ethnicity_df$ethnicity)]) +
  theme_classic() +
  labs(
    x = "Cross-validation fold",
    y = "Proportion of samples",
    fill = "Ethnicity",
    title = "Train/test ethnicity composition across held-out folds"
  )

ggsave("../output/study_held_out_ethnicity_distribution_check.pdf", width = 9, height = 5, bg = "white")
```

```{r}
sample_expression_values <- function(sce_obj, assay_name, sample_set, display_name, split_scheme, fold, n_values = 25000, transform_fn = identity) {
  vals <- as.numeric(assay(sce_obj, assay_name))
  vals <- vals[is.finite(vals)]

  if (length(vals) > n_values) {
    vals <- sample(vals, size = n_values)
  }

  tibble(
    split_scheme = split_scheme,
    fold = factor(fold, levels = seq_len(n_folds), labels = paste("Fold", seq_len(n_folds))),
    sample_set = sample_set,
    assay = display_name,
    value = transform_fn(vals)
  )
}

set.seed(split_seed)

labeltransfer_expr_qc_df <- imap_dfr(
  split_plan_list,
  function(plan, split_scheme) {
    map2_dfr(
      plan$folds,
      seq_along(plan$folds),
      ~ {
        test_i <- .x
        train_i <- setdiff(seq_len(ncol(sce_ref)), test_i)

        bind_rows(
          sample_expression_values(sce_ref[, train_i], "counts", "Training", "log1p(counts)", split_scheme, .y, transform_fn = log1p),
          sample_expression_values(sce_ref[, test_i], "counts", "Test", "log1p(counts)", split_scheme, .y, transform_fn = log1p),
          sample_expression_values(sce_query, "counts", "Query", "log1p(counts)", split_scheme, .y, transform_fn = log1p),
          sample_expression_values(sce_ref[, train_i], "logcounts", "Training", "logCPM", split_scheme, .y),
          sample_expression_values(sce_ref[, test_i], "logcounts", "Test", "logCPM", split_scheme, .y),
          sample_expression_values(sce_query, "logcounts", "Query", "logCPM", split_scheme, .y),
          sample_expression_values(sce_ref[, train_i], "logcounts_tmm", "Training", "log1pTMM", split_scheme, .y),
          sample_expression_values(sce_ref[, test_i], "logcounts_tmm", "Test", "log1pTMM", split_scheme, .y),
          sample_expression_values(sce_query, "logcounts_tmm", "Query", "log1pTMM", split_scheme, .y),
          sample_expression_values(sce_ref[, train_i], "cpm_tmm_logcounts", "Training", "logCPM(TMM)", split_scheme, .y),
          sample_expression_values(sce_ref[, test_i], "cpm_tmm_logcounts", "Test", "logCPM(TMM)", split_scheme, .y),
          sample_expression_values(sce_query, "cpm_tmm_logcounts", "Query", "logCPM(TMM)", split_scheme, .y)
        )
      }
    )
  }
)

qc_folds <- ggplot(labeltransfer_expr_qc_df %>% filter(split_scheme == "dataset_id"), aes(x = value, fill = sample_set, color = sample_set)) +
  geom_density(alpha = 0.15) +
  facet_grid(assay ~ fold, scales = "free") +
  theme_classic() +
  labs(
    x = "Expression value",
    y = "Density",
    fill = "Sample set",
    color = "Sample set",
    title = "Training/test/query expression distributions"
  )

qc_folds

ggsave("../output/gene_distribution_each_fold.pdf", width = 12, height = 8, bg = "white")
```

```{r}
lt_probs_list <- vector("list", 0)
fold_truth_pred <- vector("list", 0)

for (split_scheme in names(split_plan_list)) {
  plan <- split_plan_list[[split_scheme]]

  for (i in seq_along(plan$folds)) {
    cat(split_scheme, "fold", i, "\n")

    test_i <- plan$folds[[i]]
    train_i <- setdiff(seq_len(ncol(seu_ref)), test_i)

    seu_train <- seu_ref[, train_i]
    seu_test <- seu_ref[, test_i]

    seu_train <- NormalizeData(seu_train) %>%
      FindVariableFeatures() %>%
      ScaleData(verbose = FALSE) %>%
      RunPCA(npcs = 20, verbose = FALSE)

    feats <- VariableFeatures(seu_train)

    seu_test <- NormalizeData(seu_test) %>%
      ScaleData(features = feats, verbose = FALSE) %>%
      RunPCA(features = feats, npcs = 20, verbose = FALSE)

    anchors <- FindTransferAnchors(
      reference = seu_train,
      query = seu_test,
      dims = 1:10,
      reference.reduction = "pca",
      k.anchor = 10
    )

    predictions <- TransferData(
      anchorset = anchors,
      refdata = seu_train$ethnicity_clean,
      dims = 1:10,
      k.weight = 20
    )

    truth_clean <- seu_test$ethnicity_clean
    pred_clean <- predictions$predicted.id

    fold_truth_pred[[length(fold_truth_pred) + 1]] <- tibble(
      split_scheme = split_scheme,
      fold = i,
      truth = factor(truth_clean, levels = clean_levels),
      prediction = factor(pred_clean, levels = clean_levels),
      model = "LabelTransfer"
    )

    score_cols <- grep("^prediction\\.score\\.", colnames(predictions), value = TRUE)

    lt_prob <- predictions[, score_cols, drop = FALSE] %>%
      as_tibble() %>%
      rename_with(~ sub("^prediction\\.score\\.", "", .x)) %>%
      mutate(across(everything(), ~ as.numeric(.x))) %>%
      mutate(
        truth = factor(truth_clean, levels = clean_levels),
        fold = i,
        model = "LabelTransfer",
        split_scheme = split_scheme
      )

    missing_cols <- setdiff(clean_levels, colnames(lt_prob))
    if (length(missing_cols) > 0) {
      lt_prob[missing_cols] <- 0
    }

    lt_probs_list[[length(lt_probs_list) + 1]] <- lt_prob %>%
      dplyr::select(all_of(clean_levels), truth, fold, model, split_scheme)
  }
}

lt_model_probs <- bind_rows(lt_probs_list)
results_all <- bind_rows(fold_truth_pred)
```

```{r}
fold_metrics_lt <- results_all %>%
  group_by(split_scheme, model, fold) %>%
  summarise(
    acc = mean(truth == prediction),
    balanced_acc = bal_accuracy_vec(truth, prediction),
    macro_f1 = f_meas_vec(truth, prediction, estimator = "macro"),
    weighted_f1 = f_meas_vec(truth, prediction, estimator = "macro_weighted"),
    .groups = "drop"
  )

summary_lt <- fold_metrics_lt %>%
  group_by(split_scheme, model) %>%
  summarise(
    acc_mean = mean(acc),
    acc_sd = sd(acc),
    balanced_acc_mean = mean(balanced_acc),
    balanced_acc_sd = sd(balanced_acc),
    macro_f1_mean = mean(macro_f1),
    macro_f1_sd = sd(macro_f1),
    weighted_f1_mean = mean(weighted_f1),
    weighted_f1_sd = sd(weighted_f1),
    .groups = "drop"
  )

summary_lt

per_class_lt <- results_all %>%
  group_by(split_scheme, model, fold) %>%
  group_modify(~ {
    df <- .x
    bind_rows(lapply(clean_levels, function(cls) {
      df2 <- df %>%
        mutate(
          truth_binary = factor(if_else(truth == cls, "yes", "no"), levels = c("yes", "no")),
          pred_binary = factor(if_else(prediction == cls, "yes", "no"), levels = c("yes", "no"))
        )

      prec <- precision(df2, truth_binary, pred_binary, event_level = "first")$.estimate
      rec <- recall(df2, truth_binary, pred_binary, event_level = "first")$.estimate
      f1 <- f_meas(df2, truth_binary, pred_binary, event_level = "first")$.estimate

      tibble(
        class = cls,
        precision = ifelse(is.na(prec), 0, prec),
        recall = ifelse(is.na(rec), 0, rec),
        f1 = ifelse(is.na(f1), 0, f1),
        support = sum(df$truth == cls)
      )
    }))
  }) %>%
  ungroup() %>%
  mutate(Label = clean_to_ori_name[class]) %>%
  group_by(split_scheme, model, Label) %>%
  summarise(
    precision_mean = mean(precision),
    precision_sd = sd(precision),
    recall_mean = mean(recall),
    recall_sd = sd(recall),
    f1_mean = mean(f1),
    f1_sd = sd(f1),
    support_total = sum(support),
    .groups = "drop"
  )

per_class_lt
```

```{r}
lt_confusion_df <- results_all %>%
  dplyr::count(split_scheme, truth, prediction, name = "n") %>%
  group_by(split_scheme, truth) %>%
  mutate(prop = n / sum(n)) %>%
  ungroup() %>%
  mutate(
    truth_label = clean_to_ori_name[as.character(truth)],
    prediction_label = clean_to_ori_name[as.character(prediction)]
  )

ggplot(lt_confusion_df, aes(x = prediction_label, y = truth_label, fill = prop)) +
  geom_tile(color = "white") +
  geom_text(aes(label = n), size = 3) +
  scale_fill_viridis_c(labels = scales::percent_format(accuracy = 1)) +
  facet_wrap(~ split_scheme) +
  theme_classic() +
  labs(
    x = "Predicted ethnicity",
    y = "True ethnicity",
    fill = "Row proportion",
    title = "LabelTransfer confusion matrices across held-out folds"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
lt_roc_input <- lt_model_probs %>%
  pivot_longer(cols = all_of(clean_levels), names_to = "class", values_to = "prob") %>%
  mutate(truth_binary = factor(truth == class, levels = c(FALSE, TRUE), labels = c("no", "yes")))

lt_roc_input_ok <- lt_roc_input %>%
  group_by(split_scheme, class, fold) %>%
  filter(any(truth_binary == "yes") & any(truth_binary == "no")) %>%
  ungroup()

lt_roc_curves <- lt_roc_input_ok %>%
  group_by(split_scheme, class, fold) %>%
  roc_curve(truth_binary, prob, event_level = "second") %>%
  ungroup() %>%
  transmute(
    split_scheme = split_scheme,
    model_display = "LabelTransfer",
    ethnicity = clean_to_ori_name[class],
    fold = fold,
    fpr = 1 - specificity,
    tpr = sensitivity
  )

lt_auc_by_class <- lt_roc_input_ok %>%
  group_by(split_scheme, class, fold) %>%
  roc_auc(truth_binary, prob, event_level = "second") %>%
  ungroup() %>%
  transmute(
    split_scheme = split_scheme,
    model_display = "LabelTransfer",
    ethnicity = clean_to_ori_name[class],
    fold = fold,
    auc = .estimate
  )

lt_auc_by_class
```

```{r}
roc_py <- read_csv("../roc_points.csv") %>%
  mutate(model_display = model) %>%
  mutate(model_display = str_remove(model_display, "_S_")) %>%
  filter(model_display %in% selected_models) %>%
  transmute(
    split_scheme = split_scheme,
    model_display = model_display,
    ethnicity = ethnicity,
    fold = fold,
    fpr = fpr,
    tpr = tpr
  )

auc_py <- read_csv("../auc_by_class.csv") %>%
  mutate(model_display = model) %>%
  mutate(model_display = str_remove(model_display, "_S_")) %>%
  filter(model_display %in% selected_models) %>%
  transmute(
    split_scheme = split_scheme,
    model_display = model_display,
    ethnicity = ethnicity,
    fold = fold,
    auc = auc
  )

roc_all <- bind_rows(roc_py, lt_roc_curves) %>%
  mutate(
    split_scheme = factor(split_scheme, levels = c("dataset_id", "tissue_groups")),
    model_display = factor(model_display),
    ethnicity = factor(ethnicity)
  ) %>%
  mutate(model_display = str_remove(model_display, "_S_"))

auc_all <- bind_rows(auc_py, lt_auc_by_class) %>%
  mutate(
    split_scheme = factor(split_scheme, levels = c("dataset_id", "tissue_groups")),
    model_display = factor(model_display),
    ethnicity = factor(ethnicity)
  ) %>%
  mutate(model_display = str_remove(model_display, "_S_"))
```

```{r}
method_names <- unique(as.character(roc_all$model_display))

if (length(method_names) <= length(ning_colors)) {
  method_cols <- ning_colors[seq_along(method_names)]
} else {
  method_cols <- colorRampPalette(ning_colors)(length(method_names))
}

names(method_cols) <- method_names

fpr_grid <- seq(0, 1, by = 0.01)

interp_roc <- function(df_one, grid = fpr_grid) {
  df_one <- df_one %>%
    arrange(fpr, tpr) %>%
    group_by(fpr) %>%
    summarise(tpr = max(tpr), .groups = "drop")

  tibble(
    fpr = grid,
    tpr = approx(df_one$fpr, df_one$tpr, xout = grid, ties = "ordered", rule = 2)$y
  )
}

roc_one_line <- roc_all %>%
  group_by(split_scheme, model_display, ethnicity, fold) %>%
  group_modify(~ interp_roc(.x)) %>%
  ungroup() %>%
  group_by(split_scheme, model_display, ethnicity, fpr) %>%
  summarise(tpr = median(tpr, na.rm = TRUE), .groups = "drop")


```


```{r}
#| fig-width: 13
#| fig-height: 13
ggplot(roc_one_line %>% filter(split_scheme == "dataset_id"), 
       aes(x = fpr, y = tpr, color = model_display)) +
  geom_line(linewidth = 1) +
  facet_wrap(~model_display) +
  geom_abline(linetype = 2) +
  scale_color_manual(values = method_cols) +
  coord_equal() +
  theme_classic() +
  labs(x = "FPR", y = "TPR", color = "Method", title = "ROC curves (median across CV folds) - dataset_id held-out") +
  theme(legend.position = "none")
```

```{r}
roc_one_line_final <- roc_one_line %>% 
    filter(!grepl("CPM",model_display))

method_cols_f <- method_cols[unique(roc_one_line_final$model_display)]

roc_one_line_final <- roc_one_line_final %>% 
  mutate(model_display = str_remove(model_display,"_Log1p"))

names(method_cols_f) <- unique(roc_one_line_final$model_display)

ggplot(roc_one_line_final, 
       aes(x = fpr, y = tpr, color = model_display)) +
  geom_line(linewidth = 1, color = "gray") +
  facet_wrap(~model_display) +
  geom_abline(linetype = 2, color = "black") +
  coord_equal() +
  theme_classic() +
  labs(x = "FPR", y = "TPR", color = "Method", title = "ROC curves (median across CV folds)") +
  theme(legend.position = "none")

ggsave("../output/ROC_all.png", width = 10, height = 10, bg = "white")
```


```{r}
good_models <- c("HistGBDT_TMMScale", "HistGBDT_TMMPCA", "LabelTransfer",
                 "LinearSVM_TMMPCA", "LogReg_TMMPCA",
                 "MLP_TMMScale","RandomForest_TMMScale")

gm_cols <- c("#44545B","#DE903F","#249FD1","#BB4E49",
             "#7EAE94", "#7570AB", "#C6A24C")

names(gm_cols) <- good_models
```



```{r}
p_model_performance <- ggplot(roc_one_line_final %>% 
                                filter(split_scheme == "dataset_id") %>%
                                filter(model_display %in% good_models), aes(x = fpr, y = tpr, color = model_display)) +
  geom_line(linewidth = 1) +
  facet_wrap( ~ ethnicity, nrow = 2, scales = "free") +
  geom_abline(linetype = "dashed", color = "gray") +
  scale_color_manual(values = gm_cols) +
  theme_classic() +
  labs(x = "False-postivie rate", y = "True-positive rate", color = "Method", title = "") +
  scale_x_continuous(breaks = c(0,0.5,1)) +
  scale_y_continuous(breaks = c(0,0.5,1)) +
  theme(legend.position = "none", strip.background = element_blank(),
        strip.text = element_text(size = 12),
        axis.title = element_text(size = 12))

p_model_performance

ggsave("../output/roc_performance.pdf", width = 7, height = 5, bg = "white")
```


```{r}
p_model_performance_tho <- ggplot(roc_one_line_final %>% 
                                    filter(split_scheme == "tissue_groups") %>%
                                    filter(model_display %in% good_models), aes(x = fpr, y = tpr, color = model_display)) +
  geom_line(linewidth = 1) +
  facet_wrap( ~ ethnicity, nrow = 2, scales = "free") +
  geom_abline(linetype = "dashed", color = "gray") +
  scale_color_manual(values = gm_cols) +
  theme_classic() +
  labs(x = "False-postivie rate", y = "True-positive rate", color = "Method", title = "") +
  scale_x_continuous(breaks = c(0,0.5,1)) +
  scale_y_continuous(breaks = c(0,0.5,1)) +
  theme(legend.position = "none", strip.background = element_blank(),
        strip.text = element_text(size = 12),
        axis.title = element_text(size = 12))

p_model_performance_tho

ggsave("../output/roc_performance_tho.pdf", width = 13, height = 7, bg = "white")
```



```{r}
overall_py2 <- read_csv("../overall_performance.csv") %>%
  mutate(model = str_remove(model, "_S_")) %>%
  rename_with(~ gsub("_std$", "_sd", .x)) %>%
  mutate(model_display = model) %>%
  filter(model_display %in% selected_models)
  

pereth_py2 <- read_csv("../pereth_performance.csv") %>%
  mutate(model = str_remove(model, "_S_")) %>%
  mutate(
    model_display = model,
    Label = as.character(ethnicity)
  ) %>%
  filter(model_display %in% selected_models) %>%
  dplyr::select(-model, -ethnicity)

overall_lt2 <- summary_lt %>%
  transmute(
    split_scheme = split_scheme,
    model_display = "LabelTransfer",
    acc_mean = acc_mean,
    acc_sd = acc_sd,
    balanced_acc_mean = balanced_acc_mean,
    balanced_acc_sd = balanced_acc_sd,
    macro_f1_mean = macro_f1_mean,
    macro_f1_sd = macro_f1_sd,
    weighted_f1_mean = weighted_f1_mean,
    weighted_f1_sd = weighted_f1_sd
  )

pereth_lt2 <- per_class_lt %>%
  transmute(
    split_scheme = split_scheme,
    model_display = "LabelTransfer",
    Label = Label,
    precision_mean = precision_mean,
    precision_sd = precision_sd,
    recall_mean = recall_mean,
    recall_sd = recall_sd,
    f1_mean = f1_mean,
    f1_sd = f1_sd,
    support_total = support_total
  )

overall_all <- bind_rows(overall_py2 %>% dplyr::select(-model), overall_lt2) %>%
  mutate(
    split_scheme = factor(split_scheme, levels = c("dataset_id", "tissue_groups")),
    model_display = factor(model_display, levels = c(selected_models, "LabelTransfer"))
  )

pereth_all <- bind_rows(pereth_py2, pereth_lt2) %>%
  mutate(
    split_scheme = factor(split_scheme, levels = c("dataset_id", "tissue_groups")),
    model_display = factor(model_display, levels = c(selected_models, "LabelTransfer"))
  ) %>%
  mutate(model_display = str_remove(model_display,"_Log1p"))
```

```{r}
overall_long <- overall_all %>%
  pivot_longer(cols = ends_with("_mean"), names_to = "metric", values_to = "mean") %>%
  mutate(metric = sub("_mean$", "", metric)) %>%
  left_join(
    overall_all %>%
      pivot_longer(cols = ends_with("_sd"), names_to = "metric_sd", values_to = "sd") %>%
      mutate(metric = sub("_sd$", "", metric_sd)) %>%
      dplyr::select(split_scheme, model_display, metric, sd),
    by = c("split_scheme", "model_display", "metric")
  ) %>%
  filter(!metric %in% c("balanced_acc", "weighted_f1")) %>%
  mutate(model_display = str_remove(model_display,"_Log1p"))

ov_performance <- ggplot(overall_long %>%
         filter(split_scheme == "dataset_id",
                model_display %in% good_models), aes(x = model_display, y = mean, fill = model_display)) +
  geom_col(width = 0.75) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = gm_cols) +
  guides(fill = guide_legend(override.aes = list(shape = 15))) +
  labs(fill = "Models") +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none",
        strip.background = element_blank()) +
  labs(x = "Method", y = "Score", title = "")

ov_performance

ggsave("../output/benchmark_overall_performance.pdf", width = 7, height = 3, bg = "white")

ov_performance +
  theme(legend.position = "right") +
  guides(fill = guide_legend(ncol = 1))

ggsave("../output/benchmark_overall_performance_legend.pdf", width = 7, height = 7, bg = "white")


```

```{r}
ov_performance_tho <- ggplot(overall_long %>%
         filter(split_scheme == "tissue_groups",
                model_display %in% good_models), aes(x = model_display, y = mean, fill = model_display)) +
  geom_col(width = 0.75) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.2) +
  facet_wrap(~ metric, scales = "free_y") +
  scale_fill_manual(values = gm_cols) +
  theme_classic() +
  theme(axis.text.x = element_blank(),
        axis.ticks.x = element_blank(),
        legend.position = "none") +
  labs(x = "Method", y = "Score", title = "Overall performance - tissue held-out")

ov_performance_tho

ggsave("../output/benchmark_overall_performance_tho.pdf", width = 8, height = 6, bg = "white")
```





```{r}
pereth_long <- pereth_all %>%
  pivot_longer(cols = c(precision_mean, recall_mean, f1_mean), names_to = "metric", values_to = "mean") %>%
  mutate(metric = sub("_mean$", "", metric)) %>%
  left_join(
    pereth_all %>%
      pivot_longer(cols = c(precision_sd, recall_sd, f1_sd), names_to = "metric_sd", values_to = "sd") %>%
      mutate(metric = sub("_sd$", "", metric_sd)) %>%
      dplyr::select(split_scheme, model_display, Label, metric, sd),
    by = c("split_scheme", "model_display", "Label", "metric")
  )

pppho1 <- ggplot(pereth_long %>%
         filter(split_scheme == "dataset_id",
                model_display %in% good_models), aes(x = model_display, y = mean, fill = model_display)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), position = position_dodge(width = 0.8), width = 0.2) +
  facet_grid(Label ~ metric, scales = "free_y") +
  scale_fill_manual(values = gm_cols) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  labs(x = "Method", y = "Score", title = "Per-ethnicity performance (mean±sd across folds) - dataset held-out")

pppho1

ggsave("../output/benchmark_pereth_performance.png", width = 12, height = 10, bg = "white")
```


```{r}
pppho2 <- ggplot(pereth_long %>%
         filter(split_scheme == "tissue_groups",
                model_display %in% good_models), aes(x = model_display, y = mean, fill = model_display)) +
  geom_col(position = position_dodge(width = 0.8), width = 0.7) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), position = position_dodge(width = 0.8), width = 0.2) +
  facet_grid(Label ~ metric, scales = "free_y") +
  scale_fill_manual(values = gm_cols) +
  theme_classic() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1), legend.position = "none") +
  labs(x = "Method", y = "Score", title = "Per-ethnicity performance (mean±sd across folds) - tissue held-out")

pppho2

ggsave("../output/benchmark_pereth_performance_tho.pdf", width = 12, height = 14, bg = "white")
```

```{r}
#| fig-width: 10
#| fig-height: 15
p_model_performance_tho + ov_performance_tho + pppho2 +
  plot_layout(ncol = 1, heights = c(2,2,3)) +
  plot_annotation(tag_levels = "A")

ggsave("../output/benchmark_tho.png", width = 10, height = 15, bg = "white")
```


```{r}
pppho1 + pppho2 + plot_layout(ncol = 1)

ggsave("../output/benchmark_models_per_ethnicity.png", width =10, height = 12, bg = "white")
```



# Final prediction results

```{r}
final_predictions <- read_csv("../histgbdt_Log1pTMM_S_PCA_query_predictions.csv")

table(final_predictions$predicted_ethnicity)
```


# Markers

```{r}
library(org.Hs.eg.db)

df <- assay(sce_ref, "counts_adjusted_ethnicity") %>%
  as.data.frame() %>%
  rownames_to_column("gene") %>%
  pivot_longer(-gene, names_to = "sampleid", values_to = "counts") %>%
  left_join(colData(sce_ref) %>%
              as.data.frame() %>%
              dplyr::select("ethnicity_relabel") %>%
              rownames_to_column("sampleid")) %>%
  dplyr::select(c("ethnicity_relabel", "gene", "counts")) %>%
  group_by(ethnicity_relabel, gene) %>%
  summarise(meanc = mean(counts), .groups = "drop") %>%
  pivot_wider(names_from = "gene", values_from = "meanc") %>%
  column_to_rownames("ethnicity_relabel")

marker_gene_lookup <- tibble(gene = unique(top_markers_all$gene)) %>%
  mutate(symbol = NA_character_)

ensg_idx <- str_detect(marker_gene_lookup$gene, "^ENSG")

marker_gene_lookup$symbol[ensg_idx] <- mapIds(
  org.Hs.eg.db,
  keys = marker_gene_lookup$gene[ensg_idx],
  column = "SYMBOL",
  keytype = "ENSEMBL",
  multiVals = "first"
) %>%
  unname()

marker_gene_lookup <- marker_gene_lookup %>%
  mutate(display_gene = coalesce(symbol, gene))

marker_top5_table <- top_markers_all %>%
  left_join(marker_gene_lookup, by = "gene") %>%
  mutate(
    ethnicity_relabel = recode(
      ethnicity,
      Europeans = "European",
      HispanicDLatinAmerican = "Hispanic",
      .default = ethnicity
    )
  ) %>%
  filter(
    ethnicity_relabel %in% rownames(df),
    !is.na(display_gene),
    !str_detect(display_gene, "^ENSG")
  ) %>%
  arrange(ethnicity_relabel, desc(dist0_CI), desc(estimate)) %>%
  distinct(ethnicity_relabel, display_gene, .keep_all = TRUE) %>%
  group_by(ethnicity_relabel) %>%
  slice_head(n = 5) %>%
  ungroup() %>%
  dplyr::select(
    ethnicity_relabel, gene, display_gene,
    dist0_CI, estimate, ci_lower, ci_upper, star
  )

marker_plot_lookup <- marker_gene_lookup %>%
  distinct(gene, display_gene) %>%
  mutate(plot_label = make.unique(display_gene, sep = "_"))

marker_top5_labels <- marker_top5_table %>%
  distinct(gene, display_gene) %>%
  left_join(marker_plot_lookup, by = c("gene", "display_gene")) %>%
  pull(plot_label)

marker_genes <- marker_plot_lookup$gene
sub_mat <- as.matrix(df[, marker_genes, drop = FALSE])
colnames(sub_mat) <- marker_plot_lookup$plot_label[match(colnames(sub_mat), marker_plot_lookup$gene)]

keep_cols <- apply(sub_mat, 2, function(x) {
  all(is.finite(x)) && sd(x) > 0
})

dropped_marker_genes <- colnames(sub_mat)[!keep_cols]

sub_mat <- sub_mat[, keep_cols, drop = FALSE]
if (ncol(sub_mat) == 0) {
  stop("No finite, non-constant marker genes remained for the heatmap.")
}
sub_mat_z <- scale(sub_mat)

marker_top5_table

if (length(dropped_marker_genes) > 0) {
  tibble(dropped_marker_gene = dropped_marker_genes)
}

write_csv(marker_top5_table, "../output/top5_markers_by_ethnicity_dist0CI.csv")

cluster_cols <- my_colors

```


```{r}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(scico)

m <- sub_mat_z

genes_selected <- c("MMP8", "LGALS2", "GJB2", "IGLV8-61")

genes_selected_present <- intersect(genes_selected, colnames(m))
marker_top5_labels_present <- intersect(marker_top5_labels, colnames(m))
genes_to_mark <- unique(c(marker_top5_labels_present, genes_selected_present))
mark_idx <- which(colnames(m) %in% genes_to_mark)
mark_lbl <- colnames(m)[mark_idx]
mark_lbl_rendered <- ifelse(
  mark_lbl %in% genes_selected_present,
  paste0("<span style='color:red;'>", mark_lbl, "</span>"),
  mark_lbl
)
mark_lbl_rendered <- gt_render(mark_lbl_rendered)

if (length(setdiff(genes_selected, genes_selected_present)) > 0) {
  tibble(missing_highlight_gene = setdiff(genes_selected, genes_selected_present))
}

col_fun <- colorRamp2(
  c(min(m, na.rm = TRUE), 0, max(m, na.rm = TRUE)),
  scico(3, palette = "vik")
)

row_name_cols <- cluster_cols[rownames(m)]
row_name_cols[is.na(row_name_cols)] <- "black"

ta <- HeatmapAnnotation(
  mark = anno_mark(
    at = mark_idx,
    side = "bottom",
    labels = mark_lbl_rendered,
    labels_gp = gpar(fontsize = 9, fontface = "bold"),
    link_gp = gpar(col = "black"),
    link_width = unit(4, "mm"),
    link_height = unit(3, "mm")
  ),
  annotation_name_side = "left",
  show_annotation_name = FALSE
)

cht <- Heatmap(
  m,
  name = "Z",
  col = col_fun,
  cluster_rows = TRUE,
  cluster_columns = TRUE,
  show_row_names = TRUE,
  show_column_names = FALSE,
  row_names_gp = gpar(fontsize = 12, col = row_name_cols),
  heatmap_width  = unit(35, "cm"),
  heatmap_height = unit(8, "cm"),
  column_dend_height = unit(2.5, "cm"),
  row_dend_width = unit(2.5, "cm"),
  border = TRUE,
  bottom_annotation = ta
)

pdf("../output/Ethnicity_markerG_heatmap_full_ComplexHeatmap.pdf",
    width = 18, height = 8)
draw(cht, heatmap_legend_side = "right", annotation_legend_side = "right")
dev.off()

```



# Merge figure


```{r}
library(cowplot)
library(magick)

W   <- 800
H   <- 600  
Wt  <- 400    

p <- p_onlyhighscore

img <- image_graph(width = W, height = H, res = 96)
print(p)  # your ggplot
dev.off()

ht   <- (W - Wt) / 2
y_top <- 250        # e.g. 50px down
H2    <- 450       # bottom edge at 450px instead of 600px

coords_shrunk <- c(
  0, 0,      ht,      y_top,
  W, 0,      W - ht,  y_top,
  0, H,      0,      H2,
  W, H,      W,      H2
)

img_trap <- image_distort(
  img,
  distortion  = "perspective",
  coordinates = coords_shrunk,
  bestfit     = TRUE
)


plot(img_trap)


g <- rasterGrob(
  img_trap,
  width  = unit(1, "npc"),
  height = unit(1, "npc"),
  interpolate = TRUE
)  # keeps high fidelity :contentReference[oaicite:6]{index=6}


p_onlyhighscore_trap <- ggdraw() +
  draw_grob(
    g,
    x      = 0,
    y      = 0.2,        # leave 20% margin below
    width  = 1,
    height = 0.5         # fill 60% of canvas height :contentReference[oaicite:7]{index=7}
  ) +
  theme(plot.margin = unit(c(0,0,0,0), "pt"))



```


```{r}
p <- p_query_umap

img <- image_graph(width = W, height = H, res = 96)
print(p)  
dev.off()

ht   <- (W - Wt) / 2
y_top <- 250       
H2    <- 450      

coords_shrunk <- c(
  0, 0,      ht,      y_top,
  W, 0,      W - ht,  y_top,
  0, H,      0,      H2,
  W, H,      W,      H2
)

img_trap <- image_distort(
  img,
  distortion  = "perspective",
  coordinates = coords_shrunk,
  bestfit     = TRUE
)


plot(img_trap)


g <- rasterGrob(
  img_trap,
  width  = unit(1, "npc"),
  height = unit(1, "npc"),
  interpolate = TRUE
)  


p_query_trap <- ggdraw() +
  draw_grob(
    g,
    x      = 0,
    y      = 0.2,     
    width  = 1,
    height = 0.5        
  ) +
  theme(plot.margin = unit(c(0,0,0,0), "pt"))
```






```{r}
p_trap <- p_onlyhighscore_trap + p_query_trap + plot_layout(ncol = 1)

p_trap
```


```{r}
ht_gg <- ggplotify::as.ggplot(function() {
  draw(
    cht,
    newpage = FALSE,                  
    heatmap_legend_side = "right",
    annotation_legend_side = "right"
  )
})
```

```{r}
layout <- "
AABBDDDD
AABBDDDD
FFCCGGGG
FFCCGGGG
EEEEEEEE
EEEEEEEE
EEEEEEEE
"
```


```{r}
#| fig-width: 15
#| fig-height: 14
p_before_batch_correct + p_post_batch_correct + p_scoring_umap + p_model_performance + ht_gg + p_trap + ov_performance +
  plot_layout(design = layout, heights = c(3.5, 3.5, 5, 2, 1, 5, 1, 2),
              widths = c(3,1,1,3,3,3,1))

ggsave("../output/panel_figure.pdf", height = 14, width = 15, bg = "white")
```

