---
title: "cellNexus paper Figure 2: Dataset composition and donor coverage"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
editor: source
---

This report generates every panel of Figure 2 and then assembles the
publication figure. Keeping the data transformations, plots, and panel
descriptions together makes the analytical purpose of each panel explicit.

## Setup

```{r}
#| label: setup
#| include: false

library(cellNexus)
library(dplyr)
library(ggalluvial)
library(ggfittext)
library(ggplot2)
library(ggrepel)
library(ggsci)
library(ggridges)
library(patchwork)
library(purrr)
library(stringr)
library(tidyr)
library(viridis)

source(here("quality_control", "plot_custom_theme.R"))
source(here("quality_control", "CAQ_age_analysis_functions.R"))
source(here("quality_control", "utils.R"))
source(here("quality_control", "clean_metadata.R"))

set.seed(123)
```

## Panel A: Donor composition across technologies, tissues, and diseases

Panel A uses alluvial diagrams to show how donors from the six most
represented tissue groups are distributed across sequencing technologies and
broad disease groups. Disease groups with the three smallest donor counts are
excluded to keep low-frequency categories from obscuring the dominant
composition patterns. ScaleBio observations are excluded because that assay
was removed during metadata quality control.

```{r}
#| label: fig-panel-a
#| fig-cap: "Donor flows from technologies to tissues and from tissues to disease groups."
#| fig-width: 12
#| fig-height: 8

tissue_cols <- c(
  "Blood" = "#91AD9E",
  "Respiratory" = "#979771",
  "Cerebral Lobes" = "#B0B1B6",
  "Breast" = "#C2CEDC",
  "Skin" = "#D4BAAD",
  "Renal" = "#E1CCB1"
)

alluvial_metadata <- cell_metadata |>
  filter(assay_groups != "ScaleBio")

top_tissues_in_donors <- alluvial_metadata |>
  group_by(tissue_groups_shorten) |>
  summarise(donor_count = n_distinct(donor_id), .groups = "drop") |>
  collect() |>
  arrange(desc(donor_count)) |>
  slice_head(n = 6) |>
  pull(tissue_groups_shorten)

diseases_to_exclude <- alluvial_metadata |>
  filter(!is.na(disease_groups_coarse)) |>
  group_by(disease_groups_coarse) |>
  summarise(donor_count = n_distinct(donor_id), .groups = "drop") |>
  collect() |>
  arrange(donor_count) |>
  slice_head(n = 3) |>
  pull(disease_groups_coarse) 

tissue_disease_pairs <- alluvial_metadata |>
  filter(
    tissue_groups_shorten %in% top_tissues_in_donors,
    !disease_groups_coarse %in% diseases_to_exclude
  ) |>
  group_by(disease_groups_coarse, tissue_groups_shorten, assay_groups) |>
  summarise(donor_count = n_distinct(donor_id), .groups = "drop") |>
  collect() |>
  mutate(
    disease_groups_coarse = str_wrap(disease_groups_coarse, width = 20),
    tissue_groups_shorten = str_wrap(tissue_groups_shorten, width = 20)
  )

tissue_order <- tissue_disease_pairs |>
  summarise(total = sum(donor_count), .by = tissue_groups_shorten) |>
  arrange(desc(total)) |>
  pull(tissue_groups_shorten)

technology_order <- tissue_disease_pairs |>
  summarise(total = sum(donor_count), .by = assay_groups) |>
  arrange(desc(total)) |>
  pull(assay_groups)

disease_order <- tissue_disease_pairs |>
  summarise(total = sum(donor_count), .by = disease_groups_coarse) |>
  arrange(desc(total)) |>
  pull(disease_groups_coarse)

tissue_disease_pairs <- tissue_disease_pairs |>
  mutate(
    tissue_groups_shorten = factor(tissue_groups_shorten, levels = tissue_order),
    assay_groups = factor(assay_groups, levels = technology_order),
    disease_groups_coarse = factor(
      disease_groups_coarse,
      levels = disease_order
    ),
    tissue_text_size = case_when(
      tissue_groups_shorten == "Blood" ~ 4,
      tissue_groups_shorten == "Respiratory" ~ 3,
      tissue_groups_shorten == "Cerebral Lobes" ~ 2.5,
      tissue_groups_shorten %in% c("Breast", "Skin") ~ 2,
      tissue_groups_shorten == "Renal" ~ 1.5
    ),
    assay_text_size = case_when(
      assay_groups %in% c("10x Genomics 3", "10x Genomics 5") ~ 2.5,
      assay_groups %in% c(
        "Plate based Technologies",
        "Other Technologies",
        "Smart seq"
      ) ~ 1.5,
      assay_groups %in% c(
        "Microwell Technologies",
        "Drop based Technologies"
      ) ~ 1
    ),
    disease_text_size = case_when(
      disease_groups_coarse %in% c("Normal", "COVID-19") ~ 3,
      disease_groups_coarse %in% c(
        "Glioblastoma",
        "Lung AdenoCA",
        "Cancer",
        "Lupus"
      ) ~ 2,
      disease_groups_coarse %in% c(
        "Respiratory",
        "Metabolic/Other"
      ) ~ 1.5,
      disease_groups_coarse == "Infectious/Immune" ~ 1
    )
  ) |>
  arrange(tissue_groups_shorten)

tissue_size <- tissue_disease_pairs |>
  distinct(tissue_groups_shorten, tissue_text_size) |>
  deframe()
disease_size <- tissue_disease_pairs |>
  distinct(disease_groups_coarse, disease_text_size) |>
  deframe()
assay_size <- tissue_disease_pairs |>
  distinct(assay_groups, assay_text_size) |>
  deframe()

tissue_disease_alluvial <- tissue_disease_pairs |>
  ggplot(aes(
    axis1 = tissue_groups_shorten,
    axis2 = disease_groups_coarse,
    y = donor_count
  )) +
  geom_alluvium(
    aes(fill = tissue_groups_shorten),
    width = 0.5,
    show.legend = FALSE
  ) +
  geom_stratum(
    aes(fill = tissue_groups_shorten),
    width = 0.5,
    color = "black",
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum), size = after_stat(stratum)),
    show.legend = FALSE
  ) +
  scale_size_manual(values = c(tissue_size, disease_size)) +
  scale_x_discrete(
    limits = c("Tissue Groups", "Disease Groups"),
    expand = c(0.05, 0.05)
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = tissue_cols, na.value = NA) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = NULL) +
  theme_multipanel +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "grey95", color = NA)
  )

tissue_assay_alluvial <- tissue_disease_pairs |>
  ggplot(aes(
    axis1 = assay_groups,
    axis2 = tissue_groups_shorten,
    y = donor_count
  )) +
  geom_alluvium(
    aes(fill = tissue_groups_shorten),
    width = 0.5,
    show.legend = FALSE
  ) +
  geom_stratum(
    aes(fill = tissue_groups_shorten),
    width = 0.5,
    color = "black",
    na.rm = TRUE,
    show.legend = FALSE
  ) +
  geom_text(
    stat = "stratum",
    aes(label = after_stat(stratum), size = after_stat(stratum)),
    show.legend = FALSE
  ) +
  scale_size_manual(values = c(tissue_size, assay_size)) +
  scale_x_discrete(
    limits = c("Technologies", "Tissue Groups"),
    expand = c(0.05, 0.05)
  ) +
  scale_y_continuous(expand = c(0, 0)) +
  scale_fill_manual(values = tissue_cols, na.value = NA) +
  coord_cartesian(clip = "off") +
  labs(x = NULL, y = "Donor Count") +
  theme_multipanel +
  theme(
    legend.position = "none",
    panel.background = element_rect(fill = "grey95", color = NA)
  )

tissue_assay_alluvial + tissue_disease_alluvial
```

## Panel B: Donor demographics by tissue

Panel B summarises donor age, sex, and ethnicity for the 25 tissue groups with
the greatest donor representation. Female counts are plotted on the negative
side and male counts on the positive side; the signed log transformation
allows tissues with very different donor totals to remain visible.

```{r}
#| label: fig-panel-b
#| fig-cap: "Signed log donor counts by age, sex, ethnicity, and tissue."
#| fig-width: 8
#| fig-height: 16

donor_metadata <- cell_metadata |>
  filter(sex != "unknown") |>
  distinct(donor_id, self_reported_ethnicity, age_days, sex, tissue_groups_shorten) |>
  as_tibble() |>
  mutate(age_groups = coarse_age_bin(age_days, sex)) |>
  left_join(ethnicity_grouped)

donor_count_df <- donor_metadata |>
  count(ethnicity_groups, age_groups, sex, tissue_groups_shorten) |>
  mutate(
    age_groups = if_else(is.na(age_groups), "Unknown", age_groups),
    tissue_groups_shorten = str_replace_all(tissue_groups_shorten, "/", "/\n"),
    tissue_groups_shorten = str_wrap(tissue_groups_shorten, width = 10)
  )

tissue_order <- donor_count_df |>
  summarise(total_n = sum(n), .by = tissue_groups_shorten) |>
  arrange(desc(total_n)) |>
  pull(tissue_groups_shorten)

top_tissues <- tissue_order[seq_len(min(25, length(tissue_order)))]

demographic_plot_data <- donor_count_df |>
  filter(age_groups != "Unknown", tissue_groups_shorten %in% top_tissues) |>
  mutate(
    tissue_groups_shorten = factor(tissue_groups_shorten, levels = tissue_order),
    donor_count = if_else(sex == "female", -n, n),
    age_groups = factor(age_groups, levels = c(">60", "30-60", "<30")),
    count_sign = sign(donor_count)
  )

ethnicity_colors <- setNames(
  c(pal_npg("nrc")(7), "gray40", "gray"),
  c(
    "European",
    "East Asian",
    "Japanese",
    "South Asian",
    "African",
    "Hispanic/Latin American",
    "Native American & Pacific Islander",
    "LowConfidenceLabel",
    "Other/Unknown"
  )
)

age_sex_ethnicity_tissue_bar <- demographic_plot_data |>
  ggplot(aes(
    x = age_groups,
    y = log(abs(donor_count)) * count_sign,
    fill = ethnicity_groups
  )) +
  geom_col() +
  coord_flip() +
  geom_hline(yintercept = 0, linetype = "dashed", color = "black") +
  facet_grid(tissue_groups_shorten ~ ., switch = "y") +
  scale_fill_manual(
    values = ethnicity_colors,
    guide = guide_legend(nrow = 4)
  ) +
  labs(
    x = NULL,
    y = expression(log(abs(donor[count])) %.% Sign),
    fill = "Ethnicity"
  ) +
  theme_multipanel +
  theme(
    strip.placement = "outside",
    strip.text.y.left = element_text(
      angle = 0,
      hjust = 1,
      size = 6,
      face = "bold"
    ),
    strip.background = element_blank(),
    legend.title = element_text(size = 6),
    legend.text = element_text(size = 5),
    legend.key.height = unit(0.05, "cm"),
    legend.key.width = unit(0.05, "cm")
  )

age_sex_ethnicity_tissue_bar
```

## Panel C: Dataset scale and donor coverage

Panel C compares each dataset's donor count with its mean number of
quality-controlled (alive) cells per donor. Both axes are log-scaled because
dataset sizes span orders of magnitude. Selected extremes are labelled to
provide interpretable reference points, and colour indicates sequencing
technology.

```{r}
#| label: fig-panel-c
#| fig-cap: "Donor count versus mean alive-cell count for each dataset."
#| fig-width: 8
#| fig-height: 5

technology_colors <- c(
  "10x Genomics 3" = "#a6cee3",
  "Multi-technologies" = "#1f78b4",
  "10x Genomics 5" = "#b2df8a",
  "Other Technologies" = "#33a02c",
  "Smart seq" = "#fb9a99",
  "Microwell Technologies" = "#e31a1c",
  "TruDrop" = "#fdbf6f",
  "Drop based Technologies" = "#ff7f00",
  "Plate based Technologies" = "#cab2d6"
)

dataset_summary <- cell_metadata |>
  filter(alive) |>
  summarise(alive_cells = n(), .by = c(dataset_id, donor_id)) |>
  summarise(
    num_donor = n_distinct(donor_id),
    avg_alive_cells = mean(alive_cells),
    .by = dataset_id
  ) |>
  collect()

label_data <- dataset_summary |>
  filter(
    num_donor == max(num_donor) |
      (num_donor == min(num_donor) &
        avg_alive_cells == min(avg_alive_cells)) |
      dataset_id == "39ed7d98-676d-4b8d-9d0a-0f3b60914ead"
  ) |>
  left_join(
     cellNexus:::get_cellxgene_metadata("dataset") |>
       select(dataset_id, explorer_url),
     by = "dataset_id",
     copy = TRUE) |>
  mutate(
    author = case_when(
      str_detect(explorer_url, "d567b692-c374-4628-a508-8008f6778f22") ~
        "Kanemaru et al.",
      str_detect(explorer_url, "3faad104-2ab8-4434-816d-474d8d2641db") ~
        "Yazar et al.",
      str_detect(explorer_url, "2c820d53-cbd7-4e0a-be7a-a0ad1989a98f") ~
        "Scale Biosciences, Inc.",
      str_detect(explorer_url, "39ed7d98-676d-4b8d-9d0a-0f3b60914ead") ~
        "Smith, R. S. et al."
    )
  )

dataset_technology <- cell_metadata |>
  distinct(dataset_id, assay_groups) |>
  collect() |>
  group_by(dataset_id) |>
  mutate(
    assay_groups = if_else(
      n_distinct(assay_groups) > 1,
      "Multi-technologies",
      first(assay_groups)
    )
  ) |>
  ungroup() |>
  distinct()

label_ids <- unique(label_data$dataset_id)

alive_per_donor_scatter <- dataset_summary |>
  left_join(label_data |> select(dataset_id, author), by = "dataset_id") |>
  left_join(dataset_technology, by = "dataset_id") |>
  mutate(author = coalesce(author, "")) |>
  ggplot(aes(
    x = num_donor,
    y = avg_alive_cells,
    group = author,
    color = assay_groups
  )) +
  geom_point(size = 0.1) +
  geom_point(
    data = \(data) filter(data, dataset_id %in% label_ids),
    aes(x = num_donor, y = avg_alive_cells),
    inherit.aes = FALSE,
    color = "black",
    size = 1
  ) +
  geom_text_repel(
    data = label_data,
    aes(label = author),
    color = "black",
    size = 2,
    box.padding = 0.8,
    segment.color = "grey50",
    max.overlaps = Inf
  ) +
  scale_x_log10() +
  scale_y_log10() +
  scale_color_manual(
    values = technology_colors,
    guide = guide_legend(nrow = 3)
  ) +
  labs(
    x = "Donor Count",
    y = "Average Number of Alive Cells per Donor",
    color = "Technology"
  ) +
  theme_multipanel +
  theme(
    legend.title = element_text(size = 6),
    legend.text = element_text(size = 5),
    legend.key.height = unit(0.05, "cm"),
    legend.key.width = unit(0.05, "cm")
  )

alive_per_donor_scatter
```

## Panel D: Sample library-size distributions by technology

Panel D shows the distribution of total library size per sample for each
technology. The logarithmic x-axis accommodates the broad range of library
sizes, while the gradient encodes the log-density within each ridgeline.

```{r}
#| label: fig-panel-d
#| fig-cap: "Distribution of total library size per sample across sequencing technologies. Curves are coloured by log10-transformed peak density and ordered by median library size from top to bottom."
#| fig-width: 8
#| fig-height: 5

library_size_data <- cell_metadata |>
  summarise(
    total_library_size = sum(nCount_RNA, na.rm = TRUE),
    n_cells = n(),
    .by = c(assay_shorten, sample_id)
  ) |>
  collect()

density_peaks <- library_size_data |>
  mutate(log_total_library_size = log10(total_library_size)) |>
  group_by(assay_shorten) |>
  summarise(
    peak = list({
      density_estimate <- density(log_total_library_size, na.rm = TRUE)
      tibble(x = density_estimate$x, y = density_estimate$y) |>
        slice_max(y, n = 1, with_ties = FALSE)
    }),
    .groups = "drop"
  ) |>
  unnest(peak)

p_library_size <- library_size_data |>
  left_join(density_peaks, by = "assay_shorten") |>
  mutate(log_density_peak = log10(y)) |>
  ggplot(aes(
    x = total_library_size,
    y = reorder(assay_shorten, total_library_size, decreasing = TRUE, FUN = mean),
    fill = log_density_peak
  )) +
  geom_density_ridges_gradient(scale = 1.5, linewidth = 0.1) +
  scale_x_log10(labels = scales::label_log()) +
  scale_fill_viridis_c(
    name = expression(log[10] * "(Density Peak)"),
    option = "D",
    guide = guide_colorbar(title.position = "top", title.hjust = 0.5)
  ) +
  labs(x = "Total Library Size per Sample", y = NULL) +
  theme_multipanel +
  theme(
    legend.title = element_text(size = 5),
    legend.text = element_text(size = 4),
    legend.key.height = unit(0.2, "cm"),
    legend.key.width = unit(0.3, "cm")
  )

p_library_size
```

## Assemble Figure 2

The final layout places the paired alluvial diagrams above the dataset-scale
and library-size panels, with the demographic panel occupying the right side.

```{r}
#| label: fig-figure-2
#| fig-cap: "Figure 2: composition, demographic coverage, dataset scale, and sample library size of the cellNexus resource."
#| fig-width: 7.2
#| fig-height: 9.1

ggsave(
  filename = "tissue_disease_alluvial.pdf",
  plot = tissue_disease_alluvial,
  width = 90,
  height = 150,
  units = "mm"
)

alluvial <- (tissue_assay_alluvial + tissue_disease_alluvial) +
  plot_layout(widths = c(1, 1)) &
  theme(legend.position = "none")

left_bottom <- (alive_per_donor_scatter + p_library_size) +
  plot_layout(guides = "collect", widths = c(1, 1))

plot_left <- alluvial / plot_spacer() / left_bottom +
  plot_layout(heights = c(5, 0.01, 2))

figure_2 <- (plot_left | age_sex_ethnicity_tissue_bar) +
  plot_layout(guides = "auto", widths = c(57, 40)) &
  theme(
    plot.margin = margin(0, 0, 0, 0, "pt"),
    panel.spacing = unit(0, "cm"),
    legend.position = "bottom",
    legend.key.size = unit(0.2, "cm"),
    legend.text = element_text(margin = margin(r = 1, l = 1)),
    legend.margin = margin(t = 0, b = 0, l = -1, r = 2)
  )

# ggsave(
#   filename = here("quality_control/Figure2_dataset_content.pdf"),
#   plot = figure_2,
#   width = 183,
#   height = 230,
#   units = "mm",
#   dpi = 600
# )

figure_2
```
