---
title: "cellNexus paper Figure 3: Quality-control transformations and outcomes"
format:
  html:
    toc: true
    code-fold: true
execute:
  warning: false
  message: false
editor: source
---

Figure 3A is generated by `tissue_landscape.Qmd`. This report generates and
exports panels B--D, which document count transformations and the proportions
of cells removed by the cellNexus quality-control workflow.

## Setup

```{r}
#| label: setup
#| include: false

library(cellNexus)
library(dplyr)
library(ggplot2)
library(here)
library(patchwork)
library(purrr)
library(SummarizedExperiment)
library(tidyr)
library(zellkonverter)

source(here("quality_control", "clean_metadata.R"))
```

## Panel B: Count distributions before and after transformation

Three samples with count distributions that were difficult to classify are
shown before and after the transformations selected by the cellNexus
count transformation workflow. The comparison verifies that the inferred inverse
transformations recover plausible count-scale distributions.

```{r}
#| label: fig-panel-b
#| fig-cap: "Raw and transformed count distributions for three diagnostically challenging samples."
#| fig-width: 10
#| fig-height: 4

sample_ids <- c(
  "734b3e20708fa099e9af65400d3d30ba",
  "3d459a68fe217c54e4180fe60b7a756e",
  "dc34c7f382dd3e246b1d39ee019e6fe3"
)

get_raw_counts_per_sample <- function(sample_id) {
  counts <- zellkonverter::readH5AD(
    here("quality_control", paste0(sample_id, ".h5ad")),
    reader = "R",
    use_hdf5 = TRUE
  ) |>
    SummarizedExperiment::assay("X") |>
    colSums() |>
    as.numeric()

  tibble(
    counts = counts,
    sample_id = sample_id,
    method = "Raw"
  )
}

get_transformed_counts_per_sample <- function(sample_id) {
  counts <- cell_metadata |>
    filter(sample_id == .env$sample_id) |>
    get_single_cell_experiment() |>
    SingleCellExperiment::counts() |>
    colSums() |>
    as.numeric()

  tibble(
    counts = counts,
    sample_id = sample_id,
    method = "Transformed"
  )
}

sample_transformations <- cell_metadata |>
  filter(sample_id %in% sample_ids) |>
  distinct(sample_id, inversed_inferred_distribution) |>
  collect()

raw_counts_data <- map_dfr(sample_ids, get_raw_counts_per_sample)

transformed_counts_data <- map_dfr(
  sample_ids,
  get_transformed_counts_per_sample
) |>
  left_join(sample_transformations, by = "sample_id")

raw_density_plot <- raw_counts_data |>
  ggplot(aes(x = counts, group = sample_id)) +
  geom_density() +
  scale_x_continuous(
    labels = scales::label_number(big.mark = ",")
  ) +
  labs(
    tag = "B",
    x = "Counts",
    y = "Density"
  ) +
  theme_multipanel

transformed_density_plot <- transformed_counts_data |>
  ggplot(aes(
    x = counts,
    group = sample_id,
    linetype = inversed_inferred_distribution
  )) +
  geom_density() +
  scale_x_continuous(labels = scales::label_scientific(parse = TRUE)) +
  labs(
    x = "Transformed Counts",
    y = "Density",
    linetype = "Method"
  ) +
  theme_multipanel

panel_b <- raw_density_plot + transformed_density_plot +
  plot_layout(guides = "collect") &
  theme(legend.position = "bottom")

panel_b
```

## Panel C: Mean low-quality cell percentage by technology

For each sample, cells are classified as empty droplets, dead cells, or
doublets. Panel C reports the mean percentage of each low-quality category
across samples within every sequencing technology.

```{r}
#| label: fig-panel-c
#| fig-cap: "Mean percentage of empty droplets, dead cells, and doublets across samples for each technology."
#| fig-width: 8
#| fig-height: 5

per_sample_quality <- cell_metadata |>
  group_by(sample_id, assay_shorten) |>
  summarise(
    total_n = n(),
    empty_n = sum(empty_droplet, na.rm = TRUE),
    dead_n = sum(!empty_droplet & !alive, na.rm = TRUE),
    doublet_n = sum(
      !empty_droplet & scDblFinder.class == "doublet",
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(
    frac_empty = empty_n / total_n,
    frac_dead = dead_n / total_n,
    frac_doublet = doublet_n / total_n
  )

technology_quality <- per_sample_quality |>
  group_by(assay_shorten) |>
  summarise(
    mean_frac_empty = mean(frac_empty),
    mean_frac_dead = mean(frac_dead),
    mean_frac_doublet = mean(frac_doublet),
    .groups = "drop"
  ) |>
  mutate(
    mean_frac_good = 1 -
      mean_frac_empty -
      mean_frac_dead -
      mean_frac_doublet
  ) |>
  collect() |>
  pivot_longer(
    starts_with("mean_frac"),
    names_to = "category",
    values_to = "fraction"
  ) |>
  mutate(
    percentage = 100 * fraction,
    category = recode(
      category,
      mean_frac_good = "Alive singlet",
      mean_frac_doublet = "Doublet",
      mean_frac_dead = "Dead cell",
      mean_frac_empty = "Empty droplet"
    ),
    category = factor(
      category,
      levels = c(
        "Alive singlet",
        "Doublet",
        "Dead cell",
        "Empty droplet"
      )
    )
  )

assay_order <- technology_quality |>
  filter(category == "Alive singlet") |>
  arrange(percentage) |>
  pull(assay_shorten)

panel_c_data <- technology_quality |>
  mutate(assay_shorten = factor(assay_shorten, levels = assay_order)) |>
  filter(category != "Alive singlet")

panel_c <- panel_c_data |>
  ggplot(aes(x = assay_shorten, y = percentage, fill = category)) +
  geom_col(width = 0.7) +
  scale_fill_manual(values = c(
    "Doublet" = "#F2C14E",
    "Dead cell" = "#F26B5B",
    "Empty droplet" = "#495867"
  )) +
  labs(
    tag = "C",
    x = NULL,
    y = "Mean percentage of all droplets across samples",
    fill = NULL
  ) +
  theme_multipanel +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
    legend.position = "top"
  )

ggsave(
  here("quality_control", "Figure3C.pdf"),
  width = 120, height = 70,
  plot = panel_c, units = "mm"
)

panel_c
```

## Panel D: Doublet percentage by cell type

Panel D shows cell types whose mean sample-level doublet percentage exceeds
3%. Averaging percentages across samples prevents large samples from
dominating the cell-type comparison.

```{r}
#| label: fig-panel-d
#| fig-cap: "Cell types with a mean sample-level doublet percentage above 3%."
#| fig-width: 8
#| fig-height: 5

doublet_percentage_data <- cell_metadata |>
  filter(!is.na(cell_type_unified_ensemble)) |>
  filter(is_immune) |>
  group_by(sample_id, cell_type_unified_ensemble) |>
  summarise(
    total_cells = n(),
    doublet_cells = sum(
      scDblFinder.class == "doublet",
      na.rm = TRUE
    ),
    .groups = "drop"
  ) |>
  mutate(doublet_percentage = 100 * doublet_cells / total_cells) |>
  group_by(cell_type_unified_ensemble) |>
  summarise(
    grand_avg_doublet_pct = mean(doublet_percentage, na.rm = TRUE),
    n_samples = n(),
    .groups = "drop"
  ) |>
  filter(grand_avg_doublet_pct > 3) |>
  collect()

panel_d <- doublet_percentage_data |>
  ggplot(aes(
    x = reorder(
      cell_type_unified_ensemble,
      -grand_avg_doublet_pct
    ),
    y = grand_avg_doublet_pct
  )) +
  geom_bar(position = "stack", stat = "identity") +
  labs(
    tag = "D",
    x = "Cell type",
    y = "Avg Doublet % across Samples"
  ) +
  theme_multipanel +
  theme(
    axis.text.x = element_text(angle = 90, hjust = 1)
  )

# ggsave(here("quality_control", "Figure3D.pdf"), 
#        width = 120, height = 70, plot = panel_d, units = "mm")

panel_d
```
